# Session 1 – R setup


## Objectives

- **Explore the data‑science lifecycle.**  The *R for Data Science* introduction models data science as a cycle: you begin by **importing** and **tidying** raw data; then **understand** it through an iterative loop of **transforming**, **visualising** and **modelling**; and finally **communicate** your results.  Programming surrounds all of these steps and supports them.  Today you will learn why each component is important and how they connect.
- **Recognise the tools you need.**  To run the code in *R for Data Science* you need **R**, **RStudio**, the **tidyverse** package collection and a handful of other packages.  We will install R (from the Comprehensive R Archive Network), download RStudio (an integrated development environment), and install the tidyverse.
- **Prepare your computing environment.**  By the end of class you should have R and RStudio installed, know how to install and load packages, and be able to run simple R commands (arithmetic, vector creation, summary statistics).  We will also introduce a dataset that you will revisit throughout the semester.

## Notes

### The data‑science workflow

Data science is not a linear process.  You first **import** data from files, databases or the web; **tidy** it so that each variable is a column and each observation a row; **transform**, **visualise** and **model** your data in an iterative loop to understand patterns and relationships; and then **communicate** your findings to others.  Programming is a cross‑cutting skill that supports each of these phases.  Throughout this course, you will move back and forth between these steps rather than following them in a strict order.

### Tools and setup

You need four things to run the book’s code: **R**, **RStudio**, the **tidyverse** and some additional packages.  

- **R** is the programming language you will use.  Download the latest version  from CRAN at <https://cloud.r-project.org>.  A new major version of R is released once a year, with minor versions in between; updating regularly ensures compatibility.  
- **RStudio** is an integrated development environment (IDE) for R.  Download it from <https://posit.co/download/rstudio-desktop/>.  RStudio is updated a couple of times a year.  When you start RStudio, you will see a console pane for typing R code and an output pane for plots.
- **The tidyverse** is a collection of packages for data manipulation, visualization and programming.  To install all core tidyverse packages at once, run `install.packages("tidyverse")` in the R console.  After installation, load the tidyverse with `library(tidyverse)`; this attaches packages such as *dplyr*, *ggplot2*, *tidyr*, *readr*, *stringr*, *forcats*, *lubridate*, *purrr* and *tibble*.  You only need to install a package once, but you must load it in each new R session.
- **Other packages.**  We will occasionally use packages outside the tidyverse (e.g., *palmerpenguins*, *nycflights13*, *arrow*, *rvest*, *duckdb*).  When you encounter an error that a package is not installed, run `install.packages("package_name")` to install the package.

### Installing and testing your environment

1. **Install R and RStudio** as described above.  Accept the default installation options.  
2. **Install the tidyverse.**  Open RStudio and run the following in the console:

```{r}
#| message: false
#| warning: false
#| eval: false
install.packages("tidyverse")   # installs core tidyverse packages
library(tidyverse)               
```
   

3. **Try basic R commands.**  Use R as a calculator and practise creating vectors and computing summaries:

```{r}
2 + 2                  # arithmetic
x <- c(1, 2, 3, 5, 7)  # create a numeric vector
x * 2                  # vectorised multiplication
mean(x)                # compute the average
sum(x > 4)             # count values greater than 4 (logical vector)
```

   Notice that R performs operations element‑wise on vectors, and the assignment operator `<-` stores values.  Use descriptive variable names and indent your code neatly; we will discuss code style in a later class.


### Take‑aways

- The data‑science process is **iterative**, looping through import, tidy, transform, visualize, model and communicate.  Understanding develops as you cycle through these steps.
- To do data science in R you need the **R language**, the **RStudio IDE**, the **tidyverse** package collection and other supporting packages.  Installing and loading packages early will save time later.
- R is **vectorized**: arithmetic operations and functions operate on entire vectors.  Use the assignment operator `<-` to store results and choose descriptive variable names to write readable code.

