# Session 3 – The layered grammar of graphics

## Objectives

* **Deepen your understanding of ggplot2.** Explore the layered grammar of graphics—how aesthetic mappings, geometric objects and facets combine to build complex plots.
* **Master aesthetic mappings.** Map variables to color, shape, size and alpha correctly. Avoid mapping categorical variables to size or alpha (it implies a false ordering) and note that mapping a categorical variable to shape uses only six shapes, so additional groups are dropped.
* **Layer multiple geoms.** Add multiple geoms to a plot (e.g., points and smooth lines) and distinguish between global and local aesthetic mappings. Use the `group` aesthetic to draw separate curves for each category.
* **Use faceting and coordinate systems.** Split data into panels using `facet_wrap()` or `facet_grid()` and adjust scales. Experiment with coordinate transforms such as `coord_flip()` and `coord_polar()`.

## Notes

**Layered grammar of graphics** – Every plot can be built from layers consisting of data, aesthetic mappings, geoms, optional statistical transformations, position adjustments and a coordinate system. Building plots layer by layer allows incremental refinement.

**Aesthetic mappings** – The `aes()` function connects variables to graphical attributes. Mapping a categorical variable to color is generally safe; mapping it to shape works but only six shapes are available; mapping to size or alpha is discouraged and yields warnings.

```{r}
library(tidyverse)
library(palmerpenguins)

# Mapping species to color and shape (safe if ≤ 6 categories)
ggplot(penguins, aes(x = flipper_length_mm,
                     y = body_mass_g,
                     color = species,
                     shape = species)) +
  geom_point()

# Mapping a categorical variable to size or alpha generates warnings
ggplot(penguins, aes(x = flipper_length_mm,
                     y = body_mass_g,
                     size = species)) +
  geom_point()  # Warning: Using size for a discrete variable is not advised
```

Mappings defined in `ggplot()` apply globally, while mappings inside a geom override them for that layer.

```{r}
# Global mapping applies color to both geoms
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE)

# Local mapping overrides global color for the point layer only
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species)) +
  geom_smooth(method = "loess", se = FALSE)
```

**Layering geoms** – Different geoms (e.g., `geom_point`, `geom_smooth`, `geom_bar`) draw different types of objects. Overlaying geoms reveals multiple aspects of the data. When using geoms like `geom_smooth()`, ggplot2 automatically groups data by discrete variables; you can explicitly set `group` to control grouping.

```{r}
# Overlay scatterplot with separate curves per species
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species)) +
  geom_smooth(aes(group = species), method = "loess", se = FALSE)
```

**Facets** – Use `facet_wrap()` to create a grid of subplots for one categorical variable, and `facet_grid()` for two variables. You can allow axes to vary across facets with the `scales` argument.

```{r}
# Facet by island
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point() +
  facet_wrap(~ island)

# Facet by species and sex with free y‑axis
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  facet_grid(species ~ sex, scales = "free_y")
```

**Coordinate systems** – Transform plots using different coordinate systems to improve interpretability. For example, swap axes using `coord_flip()` or create a polar bar chart with `coord_polar()`.

```{r}
# Flip axes in a boxplot
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  coord_flip()

# Polar coordinates for a bar chart
ggplot(penguins, aes(x = species, fill = species)) +
  geom_bar() +
  coord_polar()
```



