# Session 4 – Data transformation (I): filtering, arranging, selecting and mutating

## Objectives

* **Understand the purpose of data transformation.** You rarely get data in exactly the form needed for analysis. Transformation involves creating new variables, reordering or selecting observations, and renaming columns.
* **Use dplyr verbs to manipulate rows and columns.** Learn `filter()` to subset rows, `arrange()` to reorder rows, `select()` and `rename()` to choose or rename variables, and `mutate()` to create new columns.
* **Chain operations with the pipe.** Use the pipe (`|>` in base R or `%>%` from **magrittr**) to express sequences of transformations in a readable way.

### Notes

**Key dplyr verbs** – This session covers four of the five core verbs of dplyr:

* *Pick observations by their values:* `filter()` subsets rows based on logical conditions.

  ```{r}
  library(tidyverse)
  library(palmerpenguins)
  # Filter penguins to only Adelie species on Dream island
  adelie_dream = penguins |> 
    filter(species == "Adelie", island == "Dream")
  
  adelie_dream |> head()
  ```

  `filter()` takes the data frame as its first argument and any number of logical expressions as additional arguments. It returns a new data frame and does not modify the original.

* *Reorder the rows:* `arrange()` sorts rows by one or more variables. Use `desc()` for descending order and remember that missing values are sorted to the end.

  ```{r}
  # Arrange penguins by body mass (descending) and flipper length (ascending)
  penguins_sorted = penguins |> 
    arrange(desc(body_mass_g), flipper_length_mm)
  
  penguins_sorted |> head()
  ```

* *Pick variables by their names:* `select()` quickly narrows a data frame to relevant columns. It supports helper functions like `starts_with()`, `ends_with()`, and `contains()`.

  ```{r}
  # Select species, island and body_mass_g columns
  penguins_subset = penguins |> 
    select(species, island, body_mass_g)
  
  penguins_subset |> head()
  
  # Select all columns except those from bill_length to bill_depth
  penguins_except = penguins |> 
    select(-(bill_length_mm:bill_depth_mm))
  
  penguins_except |> head()
  ```

  To rename a column without dropping others, use `rename()`.

  ```{r}
  # Rename flipper_length_mm to flipper_mm
  penguins = penguins |> 
    rename(flipper_mm = flipper_length_mm)
  
  penguins |> head()
  ```

* *Create new variables:* `mutate()` adds new columns that are functions of existing columns. You can refer to variables created earlier in the same call.

  ```{r}
  # Compute ratio of body mass to flipper length and total bill size
  penguins = penguins |> 
    mutate(
      mass_per_flipper = body_mass_g / flipper_mm,
      bill_size = bill_length_mm + bill_depth_mm
    )
  
  penguins |> head()
  ```

  If you only want to keep the new variables, use `transmute()`.

**Combining operations with the pipe** – Stringing multiple operations together can be awkward when saving intermediate objects. The pipe (`|>` or `%>%`) passes the result of one call to the next, making code easier to read.

```{r}
# Calculate average body mass by species in a single pipeline
species_summary = penguins |>
  group_by(species) |>
  summarise(
    count = n(),
    mean_mass = mean(body_mass_g, na.rm = TRUE)
  ) |>
  arrange(desc(mean_mass))

species_summary |> head()
```

The pipe should be read as “then”: group by species *then* summarise *then* arrange the result.

