# Session 5 – Data transformation (II): grouping and summarizing

## Objectives

* **Change the unit of analysis with grouping.** Use `group_by()` to organize data into subsets (groups) so that subsequent operations operate within each group rather than on the whole data set.
* **Compute summaries across groups.** Apply `summarize()` to compute summary statistics (counts, means, medians, etc.) for each group. Understand that `summarize()` returns one row for each combination of grouping variables.
* **Use helper functions.** Learn to use helpers like `n()` and `n_distinct()` within `summarize()` to count observations and distinct values.
* **Select top or bottom observations per group.** Use `slice_max()` and `slice_min()` to find the largest or smallest values within each group.
* **Enhance pipelines.** Continue to use the pipe (`|>` or `%>%`) to link `group_by()`, `summarize()` and other verbs into clear analytical workflows.

## Notes

**Grouping** – `group_by()` does not change the data itself; it changes how dplyr verbs operate. After grouping, verbs like `summarize()`, `mutate()`, and `filter()` perform their operations separately on each group.


```{r}
library(tidyverse)
library(palmerpenguins)

# Group penguins by species and island
penguins_by <- penguins |> group_by(species, island)

penguins_by
```


**Summarizing** – `summarize()` collapses multiple rows to a single row for each group. Without any grouping, it returns a single row summarizing the entire data frame.

```{r}
# Summarize mean body mass and count per species-island group
penguin_summary <- penguins |>
  group_by(species, island) |>
  summarize(
    count = n(),  # number of observations
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    distinct_years = n_distinct(year)
  )
  
  penguin_summary
```

In `summarize()`, the summary functions must return a single value. Functions such as `mean()`, `median()`, `sd()` and counts like `n()` and `n_distinct()` are typical.

**Ungrouping** – After summarizing, grouping remains; use `ungroup()` if you wish to remove grouping for subsequent operations. Alternatively, the `.groups` argument in `summarize()` can control the grouping structure of the result.

**Finding extremes within groups** – `slice_max()` and `slice_min()` extract rows with the highest or lowest values within each group.

```{r}
# Top 3 diamonds by price for each cut
top_diamonds <- diamonds |>
  group_by(cut) |>
  slice_max(order_by = price, n = 3, with_ties = FALSE)

top_diamonds
```

**Pipelines** – Continue chaining operations: group, summarize, arrange, and visualize, reading the pipe as “then”. Example:

```{r}
# Relationship between diamond carat and mean price by cut
diamond_summary <- diamonds |>
  group_by(cut) |>
  summarise(
    count = n(),
    mean_carat = mean(carat),
    mean_price = mean(price)
  ) |>
  arrange(desc(mean_price))

diamond_summary |> 
  ggplot(aes(x = cut, y = mean_price)) +
  geom_col()
```



