# Session 5: Data transformation (II): grouping and summarizing

## Objectives

- **Change the unit of analysis with grouping.** Use `group_by()` to organize data into subsets (groups) so that subsequent operations operate within each group rather than on the whole data set.
- **Compute summaries across groups.** Apply `summarize()` to compute summary statistics (counts, means, medians, and so on) for each group. Understand that `summarize()` returns one row for each combination of grouping variables.
- **Use helper functions.** Learn to use helpers like `n()` and `n_distinct()` within `summarize()` to count observations and distinct values.
- **Select top, bottom, or sampled observations per group.** Use the `slice_*()` family (`slice_max()`, `slice_min()`, `slice_head()`, `slice_tail()`, `slice_sample()`) to pull specific rows out of each group.
- **Know when grouping ends.** Understand exactly how much grouping `summarize()` removes by default, and how `ungroup()` and the `.by` argument give you explicit control.
- **Enhance pipelines.** Continue to use the pipe (`|>` or `%>%`) to link `group_by()`, `summarize()` and other verbs into clear analytical workflows.
- **Avoid the single most common grouping bug.** Recognize when leftover grouping from an earlier step silently changes what a later `filter()` or `mutate()` computes.

## Notes

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(palmerpenguins)
```

### Grouping with `group_by()`

`group_by()` does not change the data itself or reorder any rows; it only adds metadata that tells later dplyr verbs to operate separately on each group instead of on the whole data frame at once. You can see this metadata in how a grouped tibble prints.

```{r}
penguins_by <- penguins |> group_by(species, island)
penguins_by
```

Notice the `# Groups: species, island [5]` line: the data itself looks identical to `penguins`, but dplyr now knows there are five distinct species-island combinations, and every subsequent verb in the pipeline will respect that grouping until something removes it.

### Summarizing with `summarize()`

`summarize()` collapses each group down to a single row. Without any prior grouping, it collapses the entire data frame to one row. The expressions inside `summarize()` must each return a single value per group, which is why functions like `mean()`, `median()`, `sd()`, `n()` (count of rows), and `n_distinct()` (count of distinct values) show up here constantly.

```{r}
penguin_summary <- penguins |>
  group_by(species, island) |>
  summarize(
    count = n(),
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    distinct_years = n_distinct(year)
  )

penguin_summary
```

::: {.callout-note}
## Read the message dplyr prints above your output
That line reading *"summarise() has grouped output by 'species'. You can override using the .groups argument"* is not a warning to ignore; it is telling you exactly how much grouping survived the summary (see Example 5.2 for what happens when that surprises you).
:::

### Ungrouping, and the `.by` alternative

Grouping set with `group_by()` sticks around on the result until you explicitly remove it with `ungroup()`, even after a `summarize()` (see Example 5.2). If you only need grouping for a single operation, dplyr's `.by` argument lets you group and summarize (or mutate, or filter) in one step, with no lingering grouping to remember to undo afterward:

```{r}
penguins |> summarize(mean_mass = mean(body_mass_g, na.rm = TRUE), .by = species)
```

The `.by` result above has no `# Groups:` line at all, because the grouping only ever applied to that one call. This is a newer addition to dplyr (version 1.1 and later); you will see both styles, `group_by()` followed later by `ungroup()`, and one-off `.by`, in real code, and it is worth being comfortable reading either one.

### Extracting rows within groups: the `slice_*()` family

Five `slice_*()` functions pull specific rows out of each group: `slice_head()` and `slice_tail()` take the first or last `n` rows of each group as they currently appear, `slice_min()` and `slice_max()` take the rows with the smallest or largest value of some variable, and `slice_sample()` takes a random sample. All of them accept `n` for a fixed count or `prop` for a proportion of each group's rows.

```{r}
top_diamonds <- diamonds |>
  group_by(cut) |>
  slice_max(order_by = price, n = 3, with_ties = FALSE)

top_diamonds
```

::: {.callout-warning}
## `with_ties = FALSE` is not just a style preference
Without it, `slice_max()` and `slice_min()` keep *every* row tied for the cutoff position, so asking for the top 3 can hand you back 4, 5, or more rows in a group with ties. See Example 5.4 for exactly how this plays out.
:::

### Putting it together: grouped summaries in a real pipeline

Here is a small (simulated) A/B test: half of a set of website visitors saw variant A of a page, half saw variant B, and `clicked` records whether each visitor clicked a button.

```{r}
#| message: false
ab <- read_csv("data/ab_marketing.csv")

click_rates <- ab |>
  group_by(variant) |>
  summarize(
    visitors = n(),
    clicks = sum(clicked),
    click_rate = mean(clicked)
  ) |>
  arrange(desc(click_rate))

click_rates
```

```{r}
click_rates |>
  ggplot(aes(x = variant, y = click_rate)) +
  geom_col()
```

Grouping, then summarizing, then arranging, then plotting is one of the most common shapes a real analysis takes: reduce the raw data down to the handful of numbers you actually care about, then let a plot or a sorted table communicate the result.

### Fringe cases and common pitfalls

::: {.example title="Example 5.1"}
**Leftover grouping from an earlier step can silently change what a later step computes.**

Suppose you group by species early in a pipeline for one reason, then later try to find the single heaviest penguin in the entire dataset:

```{r}
# what you probably meant: the one heaviest penguin overall
penguins |> filter(body_mass_g == max(body_mass_g, na.rm = TRUE))
```

```{r}
# what you get if the data is still grouped by species from an earlier step
penguins |>
  group_by(species) |>
  select(species, island, body_mass_g) |>
  filter(body_mass_g == max(body_mass_g, na.rm = TRUE))
```

The second version returns three rows, one heaviest penguin *per species*, not the single heaviest penguin overall, because `group_by()` is still in effect and `max()` inside `filter()` is computed separately within each group. Nothing here errors or warns; the code runs perfectly and produces a perfectly reasonable-looking, wrong answer if you forgot the grouping was still active. Whenever a `filter()` or `mutate()` gives a surprising number of rows, check whether the data is still grouped from several steps earlier.
:::

::: {.example title="Example 5.2"}
**`summarize()` only removes the last level of grouping, not all of it.**

```{r}
by_species_island <- penguins |>
  group_by(species, island) |>
  summarize(n = n())
```

Grouping by two variables and then summarizing does not return an ungrouped result. It peels off only the last grouping variable (`island`) and leaves the result still grouped by the first one (`species`), which is exactly what that printed message was telling you:

```{r}
dplyr::group_vars(by_species_island)
```

If you chain another `summarize()` immediately afterward expecting it to operate on the whole table, it will instead operate per species, one more time. Setting `.groups = "drop"` inside the first `summarize()` removes all grouping in one step, which is often what you actually want at the end of a pipeline:

```{r}
penguins |>
  group_by(species, island) |>
  summarize(n = n(), .groups = "drop") |>
  dplyr::group_vars()
```
:::

::: {.example title="Example 5.3"}
**`n()` only works inside a dplyr verb, not on its own.**

```{r}
#| error: true
n()
```

`n()` (and `n_distinct()`, `cur_group()`, and a few other dplyr helpers) rely on context that only exists while dplyr is actively evaluating `mutate()`, `filter()`, `summarize()`, or `group_by()`; they have no meaning called by themselves at the console. If you see this error, you have usually copied a line out of a pipeline without its surrounding `summarize()` or `mutate()` call.
:::

::: {.example title="Example 5.4"}
**Asking for the top `n` rows per group can return more than `n` rows.**

```{r}
ties_example <- tibble(group = c("a", "a", "a"), value = c(5, 5, 3))

# asked for 1 row per group, got 2, because both 5s are tied for first place
ties_example |> group_by(group) |> slice_max(value, n = 1)

# with_ties = FALSE breaks the tie arbitrarily and returns exactly n rows
ties_example |> group_by(group) |> slice_max(value, n = 1, with_ties = FALSE)
```

If your code assumes `slice_max(..., n = 3)` always returns exactly three rows per group (for example, because you plan to immediately compute `n() == 3` or bind several groups' results into a fixed-size table), a tie will break that assumption. Decide explicitly whether ties should be kept or broken, rather than discovering the answer when a later step chokes on an unexpectedly large group.
:::

### Recap

| **Term** | **Definition** |
|-------------------|-------------------------------------------------------------|
| **`group_by()`** | Adds grouping metadata to a data frame without changing or reordering the data itself. |
| **`summarize()`** | Collapses each group to a single row using functions that return one value per group. |
| **`n()` / `n_distinct()`** | Count rows, or count distinct values, within the current group; only valid inside a dplyr verb. |
| **`.groups` argument** | Controls how much grouping `summarize()` leaves on its result: `"drop_last"` (default), `"drop"`, or `"keep"`. |
| **`ungroup()`** | Removes all grouping from a data frame explicitly. |
| **`.by`** | A per-call alternative to `group_by()` that applies only to that one verb, leaving no grouping behind afterward. |
| **`slice_max()` / `slice_min()`** | Return the rows with the largest or smallest value of a variable, within each group. |
| **`slice_head()` / `slice_tail()` / `slice_sample()`** | Return the first, last, or a random sample of rows within each group. |
| **`with_ties`** | Controls whether `slice_max()`/`slice_min()` keep every row tied for the cutoff position (default) or return exactly `n` rows. |
| **Leftover grouping** | Grouping from an earlier pipeline step that silently changes what a later `filter()` or `mutate()` computes. |

### Check your understanding

::: {.callout-note collapse="false" title="Problems"}
1. What does `group_by()` actually change about a data frame? What does it leave unchanged?
2. You run `df |> group_by(a, b) |> summarize(total = sum(x))` and then immediately run `|> summarize(grand_total = sum(total))` on the result, expecting one grand total. You get several rows instead. Explain why, and give two different fixes.
3. Why does `slice_max(x, n = 5)` sometimes return more than 5 rows for a group? How would you force it to return exactly 5?
4. What is the difference between `group_by(x) |> summarize(...)` and `summarize(..., .by = x)`? When might you prefer one over the other?
5. A pipeline groups by `region` early on for one calculation, then later in the same pipeline tries to `filter()` for the single row with the maximum `sales` across the entire dataset, but gets one row per region instead. What went wrong, and how would you fix it?
:::

::: {.callout-tip collapse="true" title="Solutions"}
1. `group_by()` adds grouping metadata that tells later dplyr verbs (`summarize()`, `mutate()`, `filter()`, and others) to operate separately within each group. It does not change any of the actual values in the data frame, add or remove rows, or reorder anything by itself.

2. By default, `summarize()` only removes the last grouping variable (`b`), leaving the result still grouped by `a`. The second `summarize()` therefore computes `sum(total)` separately within each value of `a`, producing one row per `a` instead of a single grand total. Fixing it means either adding `.groups = "drop"` to the first `summarize()`, or calling `ungroup()` on its result before summarizing again.

3. `slice_max()` keeps every row tied for the cutoff position by default, so if there is a tie at the 5th-largest value, you can get 6 or more rows back for that group. Adding `with_ties = FALSE` forces it to return exactly `n` rows by breaking ties arbitrarily.

4. `group_by(x) |> summarize(...)` leaves the grouping (minus one level) attached to the result, so it persists into whatever comes next in the pipeline unless you remove it. `summarize(..., .by = x)` groups only for that single call and leaves no grouping behind afterward. `.by` is convenient exactly when you want grouped behavior for one step without having to remember to `ungroup()` later.

5. The `group_by(region)` from earlier in the pipeline was never removed (with `ungroup()`, `.groups = "drop"`, or by switching to `.by`), so the later `filter(sales == max(sales))` computed the maximum separately within each region instead of across the whole dataset. Removing the grouping before that `filter()` (or restructuring the earlier step to use `.by` instead of `group_by()`) fixes it.
:::
