# Session 6 – Data import: reading CSV & flat files

## Objectives

* **Read your own data into R.**  Up to now we have worked with datasets that come bundled with packages.  In practice you will often need to import data from your own files.  This session introduces the `readr` functions for reading plain‑text rectangular data (CSV, TSV and other delimited formats).  You will learn how to specify file paths and handle column names, types and missing values.
* **Control column names, missing values and types.**  When you run `read_csv()` it prints the number of rows and columns read and a summary of the column specification.  You will learn how to rename variables, skip header lines, define which strings should be treated as missing, and override the type guessing heuristic that readr uses.
* **Read multiple files and combine them.**  Real projects often involve multiple data files (for example, one file per month).  You can pass a vector of file paths to `read_csv()` and use the `id` argument to record the source of each observation.  We will practice finding files with `list.files()`, reading them into a single tibble and combining them using `bind_rows()`.

## Notes

### Why import data?

Working with data included in R packages is convenient when you are learning, but eventually you need to apply the tools to your own data.  *R for Data Science* notes that this chapter focuses on **reading plain‑text rectangular files** and provides practical advice for handling **column names, types and missing data**.  The goal is to help you get your data into a tidy tibble so that you can immediately start transforming and visualizing it.

### Reading CSV files with `readr`

The `readr` package is part of the tidyverse and provides fast functions for reading delimited files.  The most common function is `read_csv()`, which expects a path to a comma‑separated file.  When you read a file, readr prints a message showing the number of rows and columns, the delimiter used and the column specification.  This message includes the guessed type for each column and can be silenced with `show_col_types = FALSE`.

```{r}
library(tidyverse)

# Example: reading a simple CSV file from a URL
students = read_csv("https://pos.it/r4ds-students-csv")
students

# Suppress the column spec message
students_quiet = read_csv("https://pos.it/r4ds-students-csv", show_col_types = FALSE)
```

Use the `file` argument to read files stored locally.  It is good practice to keep data in a dedicated `data/` folder within your project and refer to it using relative paths.  If your file has a header row containing the names of the columns (as most CSVs do), readr uses it automatically.  If not, set `col_names = FALSE` and optionally provide a vector of names via `col_names = c("...", ...)`.

### Handling missing values and non‑syntactic names

A CSV file does not encode missing values explicitly, so readr treats **empty fields** as `NA`.  In practice, missing values are often recorded with sentinel strings such as `"N/A"` or `"."`.  You can tell `read_csv()` which strings should be considered missing using the `na` argument.  For example, in the `students.csv` file the string `"N/A"` is used to mark missing foods.  By specifying `na = c("N/A", "")`, readr converts those strings to `NA`.

```{r}
students2 = read_csv(
  "https://pos.it/r4ds-students-csv",
  na = c("N/A", "")
)
students2
```

Sometimes column names contain spaces or other characters that make them **non‑syntactic** in R.  In that case they are surrounded by backticks in the tibble, and you must use backticks to refer to them.  A simple solution is to rename the columns after import using `rename()` or the `janitor::clean_names()` helper (the latter converts names to `snake_case`).

### How readr guesses column types

Because a CSV file does not contain type information, readr **guesses the type of each column**.  It samples up to 1,000 values from across the file and asks: *does the column contain only logical values?* *Only numbers?* *Does it match ISO8601 date format?* If none of those tests succeed, readr assumes the column is a string.  This heuristic works well for clean datasets but can fail when there are unexpected values (e.g., a period `.` to represent a missing number).

When type guessing fails you can provide your own **column specification** via the `col_types` argument.  This argument is a named list or a `cols()` specification where each name matches a column and each value is a type function (e.g., `col_double()`, `col_integer()`, `col_character()`, `col_date()`, `col_datetime()`).  For example, if a numeric ID is misread as a number but should be a character (because you never intend to sum it), specify `col_types = list(student_id = col_character())`.  readr provides nine column types including 

- logicals
- doubles
- integers
- characters
- factors
- dates
- date‑time
- permissive numbers
- skipped columns

```{r}
# A small CSV with a custom missing value "."
simple_csv = "x\n10\n.\n20\n30"

# Default guess treats the period as a string
df_default = read_csv(simple_csv)
df_default

# Override the type to double and inspect problems
df_num = read_csv(simple_csv, col_types = list(x = col_double()))
problems(df_num)  # shows where parsing failed

# Tell readr to treat "." as NA so guessing succeeds
df_fixed = read_csv(simple_csv, na = ".")
df_fixed
```

If you have many columns with the same type, you can set a **default type** using `cols(.default = col_character())`.  To read only a subset of columns, use `cols_only()` with the columns you want.

### Reading multiple files and combining them

In many projects you receive data split across multiple files—perhaps one file per month or per site.  Instead of reading each file separately and then binding the results, you can pass a **vector of file paths** to `read_csv()`.  readr will read all of them and stack the rows together.  The optional `id` argument adds a column that records the file each row came from.

```{r}
#| eval: false

# Suppose you have three monthly sales files stored in data/01-sales.csv, 02-sales.csv, 03-sales.csv
sales_files = c("data/01-sales.csv", "data/02-sales.csv", "data/03-sales.csv")

# Read and combine, adding a 'file' column to identify the source
sales = read_csv(sales_files, id = "file")
```

You often don’t know the names of all the files ahead of time.  Use `list.files()` to find files whose names match a pattern (e.g., `"sales\\.csv$"`) and set `full.names = TRUE` so the full path is returned.

```{r}
#| eval: false

# Find all CSV files ending with 'sales.csv' in the data directory
sales_files = list.files("data", pattern = "sales\\.csv$", full.names = TRUE)
sales_files
sales = read_csv(sales_files, id = "file")
```

Once you have imported multiple files you can combine them explicitly with `dplyr::bind_rows()` or `bind_cols()`, depending on whether you want to stack rows or columns.  `bind_rows()` requires identical column names, so pay attention to names and types when reading.

### Writing data back to disk

`readr` also provides `write_csv()` and `write_tsv()` for saving tibbles to disk.  Remember that when you write to CSV the column specification is lost—you must regenerate types when reading the file again.  For intermediate results consider using `write_rds()` and `read_rds()`, which store R objects in a binary format preserving types.

<!-- ## In‑class activity (≈60 min) -->

<!-- Today’s activity gives you hands‑on practice importing and combining data.  Work individually or in small groups.  A shared dataset (e.g., **nycflights13** or **palmerpenguins**) will be split into several CSV files ahead of time.  Download the files from the course site or find them in your project’s `data/` folder. -->

<!-- 1. **Inspect the files (10 min)** – Each student chooses one of the CSV files and opens it in a text editor or uses `read_csv()` with `n_max = 5` to glimpse the first few rows.  Note the column names and any potential issues: missing values recorded as strings (e.g., `"."` or `"NA"`), inconsistent capitalization, or unexpected characters. -->

<!-- 2. **Import with `read_csv()` (15 min)** – Use `read_csv()` to import your file.  If you see a warning about parsing problems, call `problems()` to identify the row and column causing trouble.  Adjust the `na` argument or supply a `col_types` list to fix the issue.  Discuss with your partner which types were guessed correctly and which ones needed manual specification.  Try adding `show_col_types = FALSE` to reduce clutter in your console. -->

<!-- 3. **Combine files (20 min)** – As a group, collect the names of all the CSV files and store them in a character vector.  Use `list.files()` with a pattern if the files follow a naming convention (e.g., `*_penguins.csv`).  Read all files at once using `read_csv(file_vector, id = "source")`.  Inspect the resulting tibble—does it include a new `source` column?  Alternatively, read the files one by one and combine them using `bind_rows()`. -->

<!-- 4. **Explore the combined data (15 min)** – Verify that the combined dataset has the expected number of rows and that the columns have consistent types.  Use `summarize()` to compute simple summaries (e.g., number of observations per file, mean of a numeric variable).  Discuss strategies for detecting import errors: inconsistent column names, mismatched types or unusual missing values.  Reflect on how the import step lays the groundwork for the tidy data work you will do in the next session. -->

<!-- This activity will form the basis for the tidy‑data exercises in Session 7.  Save your cleaned and combined tibble to your project folder using `write_csv()` or `write_rds()` so you can use it again. -->

