# Session 7 – Tidy data & pivoting

## Objectives

* **Define tidy data.**  Understand the three rules that make a dataset tidy: each **variable** must live in its own column, each **observation** must occupy its own row, and each **value** must appear in a single cell.  Appreciate why a consistent data structure makes it easier to learn and use tidyverse tools.
* **Lengthen data with `pivot_longer()`.**  Use `pivot_longer()` from the *tidyr* package to reshape untidy datasets by gathering column names into a new variable and their values into another variable.  Learn how to select columns to pivot, specify the names of the new variables and optionally drop missing values.
* **Handle multiple variables in column names.**  Recognize when column headers encode multiple pieces of information (e.g., method, gender and age) and use the `names_to`/`names_sep` arguments of `pivot_longer()` to split them into separate variables.
* **Widen data with `pivot_wider()`.**  Use `pivot_wider()` to spread rows into columns when each observation is represented across multiple rows.  Learn how to choose the `names_from`, `values_from` and `id_cols` arguments so that each row uniquely identifies an observation.

## Notes

### What is tidy data?

Tidy data is a **standard way to organize your datasets** so that they work naturally with the tidyverse.  In tidy data:

1. **Each variable is a column, each column is a variable.**  A dataset like `table1` from *R for Data Science* has one column for each variable and is easiest to work with.
2. **Each observation is a row.**  Each row of a tidy data frame corresponds to one observation of all variables.
3. **Each value is a cell.**  Every cell contains a single value for one variable in one observation.

The pay‑off for tidying data is twofold.  First, adopting a single consistent structure makes it easier to learn a suite of tools because they all assume the same underlying format.  Second, placing variables in columns allows R’s vectorised functions to operate naturally.

### Lengthening data with `pivot_longer()`

Most real datasets aren’t tidy because they are organised for data entry or reporting rather than analysis.  The **`pivot_longer()`** function lengthens data by gathering a set of columns into key–value pairs.  Its most important arguments are:

* `cols`: which columns to pivot.  You can specify them explicitly (`bp1:bp2`) or using tidyselect helpers such as `starts_with()`.
* `names_to`: the name of the new variable created from column names.  For example, converting week columns (`wk1`, `wk2`, …) into a variable called `week`.
* `values_to`: the name of the variable that will hold the values from the pivoted columns.
* `values_drop_na`: set to `TRUE` to drop rows where all the pivoted columns contain `NA`.

Here is a simple example using the built‑in `billboard` dataset, which records weekly Billboard chart positions.  Each row is a song and each `wk?` column gives its rank in that week.  To tidy this dataset, we gather the week columns into a new `week` variable and the ranks into a `rank` variable:

```{r}
#| message: false
#| warning: false
library(tidyverse)
billboard_long <- billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) |>
  mutate(week = parse_number(week))
billboard_long |> head()
```

This call transforms the 317 × 79 `billboard` tibble into a 5307 × 5 tibble with one row per song–week combination.

When column names contain **multiple pieces of information**, you can split them into several new variables by supplying a vector of names and a separator.  For example, the `who2` dataset records TB cases with column names like `sp_m_014`, which combine the method (`sp`), gender (`m`) and age range (`014`).  The following call extracts those parts into separate variables:

```{r}
who2_long <- who2 |>
  pivot_longer(
    cols = !(country:year),
    names_to = c("diagnosis", "gender", "age"),
    names_sep = "_",
    values_to = "count"
  )
who2_long |> head()
```

The **`names_sep`** argument splits each original column name at the underscore and assigns the pieces to the new variables (`diagnosis`, `gender`, `age`).  The **`values_to`** argument stores the counts.

### Widening data with `pivot_wider()`

Sometimes a single observation is spread across multiple rows, and you need to widen the data by creating new columns.  The function **`pivot_wider()`** increases the number of columns and decreases the number of rows.  It is particularly useful when you have long data with a variable that identifies the type of measurement and another variable with the corresponding value.

The key arguments are:

* `names_from`: the column whose unique values will become new column names.
* `values_from`: the column containing the values to fill those new columns.
* `id_cols`: optional columns that uniquely identify each row; if you omit this, `pivot_wider()` will attempt to infer them but may produce duplicate rows.

As an example, the `cms_patient_experience` dataset from the Centers for Medicare & Medicaid Services records multiple performance measures for each healthcare organization.  Each organization appears on multiple rows, one per measure.  To put each organization on its own row with separate columns for each measure code, use `pivot_wider()`:

```{r}
#| message: false
#| warning: false
cms_wide <- cms_patient_experience |>
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
cms_wide |> head()
```

By specifying the identifier columns (`org_pac_id` and `org_nm`), `pivot_wider()` ensures that each organisation occupies a single row and creates a column for each measure code.  If you omit `id_cols`, you may see duplicate rows because `pivot_wider()` cannot uniquely identify observations on its own.

### Key take‑aways

* **Tidy data has a consistent structure:** variables in columns, observations in rows and values in cells.  This organization simplifies analysis and leverages R’s vectorization.
* **`pivot_longer()` gathers columns into key–value pairs**.  Use `cols` to select columns to gather, `names_to` and `values_to` to name the new variables, and `values_drop_na` to remove structural missing values.
* **Split multiple pieces of information** encoded in column names with `names_to` and `names_sep`.
* **`pivot_wider()` spreads values across new columns**, reversing a `pivot_longer()` when observations are spread across rows.  Use `names_from`, `values_from` and (optionally) `id_cols` to control the reshaping.
* **Practice on real data.**  Tidy datasets are not always ready-made; most analyses require some tidying.  Being proficient with pivoting functions allows you to structure your data before visualizing, transforming or modelling.



