# Session 8 – Variable types I: logical & numeric vectors

## Objectives

* **Distinguish logical and numeric vectors.**  Logical vectors contain only `TRUE`, `FALSE` or `NA`, whereas numeric vectors contain integers or doubles.  Learn to convert strings to numbers with `parse_double()` and `parse_number()`.
* **Create and combine logical vectors.**  Use comparison operators (`<`, `<=`, `>`, `>=`, `!=`, `==`) to generate logical vectors and combine them with `&`, `|`, `!` and `xor()`.  Avoid short‑circuiting operators (`&&`, `||`) inside data‑masking verbs.  Understand how missing values propagate through comparisons and Boolean operations.
* **Summarize logical vectors.**  Collapse logical vectors with `any()` and `all()` or coerce them to numeric and use `sum()`/`mean()` to count or compute the proportion of `TRUE`s.  Detect missing values with `is.na()`.
* **Perform numeric operations and transformations.**  Recognize that R recycles shorter vectors in arithmetic.  Use `pmin()`/`pmax()` for element‑wise minima and maxima, `%/%` and `%%` for modular arithmetic, logarithms for rescaling, `round()`, `floor()` and `ceiling()` for rounding, and `cut()` to bin numeric data.
* **Compute numeric summaries.**  Compare the mean and median—the mean is sensitive to extreme values while the median is robust.  Use quantiles to summarize tails and spread measures such as the standard deviation and interquartile range.

## Notes

### Logical vectors and comparisons

Logical vectors are created by comparing values.  They can take on three values: `TRUE`, `FALSE` or `NA`.  Comparison operators return a logical vector with the same length as the input.  For example:

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(nycflights13)

flights |>
  mutate(late_dep = dep_delay > 30,      # TRUE if departure delay > 30 min
         early_arr = arr_delay < 0) |>   # TRUE if arrival was early
  select(year:day, dep_delay, arr_delay, late_dep, early_arr) |>
  head(5)
```

Be cautious when comparing floating‑point numbers.  Tiny rounding errors mean that equality tests may fail.  To check approximate equality use `dplyr::near()`.

### Boolean algebra and missing values

Combine logical vectors using Boolean algebra.  In R, `&` means **and**, `|` means **or**, `!` means **not**, and `xor()` is exclusive‑or.  The short‑circuiting operators `&&` and `||` return a single `TRUE`/`FALSE` and should not be used inside functions like `filter()`.

Missing values are contagious: any comparison involving `NA` yields `NA`.  In Boolean operations, `NA | TRUE` evaluates to `TRUE` but `NA | FALSE` remains `NA`.  Use `is.na()` to test for missing values:

```{r}
flights |> 
  summarise(
    n_missing_dep = sum(is.na(dep_time)),
    n_missing_arr = sum(is.na(arr_time))
  )
```

When combining conditions, remember the order of operations.  To filter flights departing in November **or** December, write `month == 11 | month == 12` or use `%in%`; writing `month == 11 | 12` mistakenly recycles the vector `c(11, 12)`.

### Summarising logical vectors

Logical summaries collapse a logical vector to a single value.  `any(x)` returns `TRUE` if **any** element of `x` is `TRUE`, while `all(x)` returns `TRUE` only if **every** element is `TRUE`.  Both functions accept `na.rm = TRUE` to ignore missing values.  For example, to check if all flights on a given day departed within an hour or if any flights arrived more than five hours late:

```{r}
flights |>
  group_by(year, month, day) |>
  summarise(
    all_dep_within_hour = all(dep_delay <= 60, na.rm = TRUE),
    any_long_arr_delay  = any(arr_delay >= 300, na.rm = TRUE),
    .groups = "drop"
  )
```

Since logical vectors coerce to numeric (`TRUE` = 1, `FALSE` = 0), `sum()` counts the number of `TRUE` values and `mean()` computes their proportion.  For instance, the proportion of flights departing within an hour and the number with very long arrival delays can be calculated by:

```{r}
flights |>
  group_by(year, month, day) |>
  summarise(
    prop_on_time_dep    = mean(dep_delay <= 60, na.rm = TRUE),
    count_long_arr_delay = sum(arr_delay >= 300, na.rm = TRUE),
    .groups = "drop"
  )
```

Logical vectors also enable **inline subsetting**.  Rather than filtering the entire data frame, you can subset a single vector with a logical condition (e.g., `arr_delay[arr_delay > 0]`) to compute summaries only on values meeting a criterion.

### Numeric vectors: making numbers and parsing

Numeric vectors may be integers or doubles.  When numbers are stored as text (e.g., `"$1,234"` or `"59%"`), use `readr::parse_double()` to convert simple numeric strings and `readr::parse_number()` to strip extraneous characters like currency symbols or percent signs.

### Vectorized arithmetic and recycling

R performs arithmetic element‑wise and recycles the shorter vector to match the length of the longer one.  Recycling single numbers (e.g., `x / 5`) is convenient, but recycling longer vectors can produce warnings or silent errors.  Avoid using `==` with vectors of unequal length; instead use `%in%` when matching multiple values.

Element‑wise minima and maxima are computed with `pmin()` and `pmax()`.  Modular arithmetic operators `%/%` and `%%` perform integer division and find remainders.  For example, unpacking a four‑digit departure time into hours and minutes:

```{r}
flights |>
  mutate(
    sched_hour   = sched_dep_time %/% 100,
    sched_minute = sched_dep_time %% 100
  ) |>
  select(sched_dep_time, sched_hour, sched_minute) |>
  head()
```

### Numeric transformations

To handle wide ranges of values, apply log transformations.  `log2()` and `log10()` are easier to interpret; for example, a difference of 1 on the log2 scale corresponds to doubling or halving the original value.  Rounding functions include `round()` (which uses banker’s rounding), `floor()` and `ceiling()`.  To round to arbitrary multiples, scale the vector, round, and then scale back.  Use `cut()` to convert a continuous variable into categories by specifying break points and optional labels.

```{r}
x = c(1, 2, 5, 10, 15, 20)
cut(x, breaks = c(0, 5, 10, 20), labels = c("small", "medium", "large"))
```

### Numeric summaries

Summarizing numeric vectors involves measures of center and spread.  The **mean** is sensitive to extreme values, while the **median** is more robust.  The median daily departure delay is always smaller than the mean because flights can be hours late but rarely leave hours early.  Quantiles generalize the median; for example, the 95th percentile ignores the most extreme 5% of values.  Measures of spread include the **standard deviation** and the **interquartile range (IQR)**, the difference between the 75th and 25th percentiles.

```{r}
flights |>
  summarise(
    mean_arr_delay    = mean(arr_delay, na.rm = TRUE),
    median_arr_delay  = median(arr_delay, na.rm = TRUE),
    q95_arr_delay     = quantile(arr_delay, 0.95, na.rm = TRUE),
    sd_arr_delay      = sd(arr_delay, na.rm = TRUE),
    iqr_arr_delay     = IQR(arr_delay, na.rm = TRUE)
  )
```

### Key take‑aways

* Logical vectors arise from comparisons and contain only `TRUE`, `FALSE` and `NA`.  Combine conditions with Boolean operators and handle missing values carefully.
* Summarize logical vectors with `any()`, `all()`, `sum()` and `mean()` to answer questions like “Were all flights on time?” or “What fraction of flights left within an hour?”.
* Be mindful of type when working with numbers.  Parse numeric strings properly and avoid unintended recycling in arithmetic and comparisons.
* Use numeric transformations—element‑wise minima/maxima, modular arithmetic, logs, rounding and binning—to prepare variables for analysis.
* Choose appropriate summaries: mean vs. median for center, quantiles for tails, and standard deviation or IQR for spread.
