# Session 9 – Variable types II: strings & regular expressions

## Objectives

* **Create and manipulate strings.**  Recall that strings are sequences of characters defined by quotes.  In R you can use single `'` or double `"` quotes to create strings.  The R4DS text recommends using double quotes by default and switching to single quotes only when the string itself contains quotes.  You’ll learn how to escape special characters like backslashes and newlines, how to measure string length with `str_length()`, and how to extract or replace substrings with `str_sub()`.
* **Combine and format strings.**  Concatenate strings using `str_c()` (stringr’s equivalent to `paste0()`), and build templated strings with `str_glue()`.  Understand how these functions handle missing values and learn to collapse vectors of strings with `str_flatten()`.
* **Understand regular expressions.**  A **regular expression (regex)** is a concise language for describing patterns within strings.  You’ll learn the difference between **literal** characters and **metacharacters**, how quantifiers like `?`, `*` and `+` control repetition, and how to use character classes (`[...]`), alternation (`|`) and anchors (`^`, `$`).
* **Match, extract and replace patterns.**  Use `str_detect()` to test whether strings match a regex, `str_count()` to count the number of matches, `str_extract()` (and `str_extract_all()`) to pull out matching substrings, `str_replace()`/`str_replace_all()` to substitute patterns, and `str_split()` to separate strings at delimiters.  Pair these functions with dplyr verbs to filter, mutate or summarize character variables.
* **Tidy multiple variables encoded in one string.**  Recognise when a single string stores several pieces of information and use `stringr` plus tidyr functions such as `str_split()` with `unnest_longer()` or `separate()` to unpack them.  This sets the stage for the next session on factors.

## Notes

### Creating strings: quotes, escapes & length

Strings can be created with single or double quotes.  Double quotes are preferred by convention—use single quotes only when the string itself contains double quotes to avoid escaping.  Special characters (backslash, newline, tab) require escaping with `\`, `\n` or `\t`.  For example:

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(stringr)

# A string containing both single and double quotes
quote_example =  "He said, 'R is great!'"
quote_example

# A string with a backslash and a newline
path =  "C:\\Users\\Data\\mydata\n"
path
```

The *length* of a string (number of characters) is computed with `str_length()`.  Note that `str_length()` counts human‑visible characters, not bytes or graphemes—R treats each character as one unit regardless of encoding:

```{r}
str_length(c("", "abc", "😊"))
```

Use `str_sub()` to extract or replace substrings by position.  Positive indices count from the start and negative indices count from the end.  The function is vectorised, so you can operate on a vector of strings at once:

```{r}
fruit =  c("banana", "apple", "pear")
str_sub(fruit, 1, 3)      # first three letters
str_sub(fruit, -3, -1)    # last three letters
str_sub(fruit, 1, 1) =  str_to_upper(str_sub(fruit, 1, 1))  # capitalize first letter
fruit
```

### Combining and formatting strings

`str_c()` concatenates multiple strings together and returns a character vector.  It’s similar to `paste0()` but has a more consistent interface and works seamlessly with dplyr verbs.  You can supply a `sep` argument to insert a separator and a `collapse` argument to collapse the result into a single string:

```{r}
species =  c("Adelie", "Chinstrap", "Gentoo")
n =  c(3, 2, 1)
str_c(n, species, sep = " ", collapse = "; ")
```

When concatenating across rows in a tibble, wrap `str_c()` inside `mutate()`.  If any of the components are `NA`, the result will be `NA`.  Use `coalesce()` to replace missing values with an empty string before concatenation:

```{r}
library(dplyr)
df =  tibble(first = c("John", NA, "Jenny"),
             last  = c("Smith", "Nguyen", NA))
df |> mutate(full = str_c(coalesce(first, ""), coalesce(last, ""), sep = " "))
```

`str_glue()` from the *glue* package (loaded automatically with tidyverse) creates templated strings.  Inside curly braces `{}`, you can insert arbitrary R expressions.  Missing values are converted to the string "NA":

```{r}
library(glue)
df |> mutate(message = str_glue("Hello {first} {last}!"))
```

If you need to combine many strings into one, use `str_flatten()` rather than repeatedly adding separators.  It collapses a character vector into a single string with a given delimiter:

```{r}
letters =  c("a", "b", "c")
str_flatten(letters, collapse = ", ")
```

### Regular expressions: pattern basics

A **regular expression (regex)** is a concise language for describing sets of strings.  In regex, most letters and numbers match themselves (are *literals*), but certain characters have special meaning (*metacharacters*).  Important metacharacters include:

* `.` matches any character except a newline.
* Quantifiers control repetition: `?` matches zero or one occurrence; `*` matches zero or more; `+` matches one or more.
* Character classes like `[aeiou]` match any of the enclosed characters.  A caret (`^`) at the start of a class negates it: `[^0-9]` matches any non‑digit.
* Alternation `|` matches one of several possibilities; for example, `apple|pear` matches either word.
* Anchors `^` and `$` match the start and end of a string, respectively.

Use `regex()` to build regexes with settings such as case‑insensitive matching (`regex(pattern, ignore_case = TRUE)`).  The `stringr` package provides helper functions like `str_view()` and `str_view_all()` to visualize matches, but we’ll focus on functions that return useful vectors.

### Detecting, counting and extracting patterns

`str_detect()` tests whether each element of a character vector matches a regex and returns a logical vector.  This pairs naturally with `filter()` to select rows matching a pattern.  `str_count()` counts the number of times a pattern appears within each string.  `str_extract()` returns the first match, and `str_extract_all()` returns **all** matches as a list.  `str_replace()` and `str_replace_all()` replace the first or all matches with a new string.  Finally, `str_split()` splits strings at matches of a regex and returns a list of pieces.

Let’s illustrate with flight numbers from the `nycflights13` dataset.  Airlines in the United States identify a flight by a two‑letter *carrier* code followed by a *flight number*.  We can build a flight ID string and then extract the numeric part:

```{r}
#| message: false
library(nycflights13)
flights_small =  flights |> select(carrier, flight) |> sample_n(5)

# Combine carrier and flight into a single string
flights_small =  flights_small |> mutate(flight_id = str_c(carrier, flight, sep = "-"))

# Detect whether the flight number has exactly 3 digits
flights_small |> mutate(is_three_digit = str_detect(flight_id, "-\\d{3}$"))

# Extract the numeric part of the flight ID
flights_small |> mutate(number_only = str_extract(flight_id, "\\d+"))
```

Notice that `\d` in a regex matches any digit (short for `[0-9]`), so `\d+` extracts one or more consecutive digits.  We had to escape the backslash in the string to make sure it reached the regular‑expression engine.

`str_replace_all()` lets you perform substitutions.  For example, you can strip all non‑digit characters from a string using the negated class `[^0-9]`:

```{r}
messy =  c("(202) 555-0198", "+1-303-555-1212")
str_replace_all(messy, "[^0-9]", "")
```

Finally, `str_split()` (or `str_split_fixed()`) breaks a string into pieces at matches of a pattern.  If you know your delimiter, you can also use tidyr’s `separate()` or `separate_wider_delim()` to split a column into multiple variables.  For example, suppose we have a vector of names in “LAST, First Middle” format:

```{r}
names =  c("SMITH, John A.", "O'NEILL, Anne", "Lee, Chen")
# Split at the comma and whitespace
df_names =  tibble(full = names) |> separate(full, into = c("last", "rest"), sep = ",\\s*")
df_names
```

### Working with messy strings in tibbles

Character columns often encode multiple variables in one field.  To tidy them, combine `stringr` and tidyr.  As an example, consider the built‑in `who2` dataset where column names like `sp_m_014` describe the diagnosis, gender and age group.  We can pivot the data longer and then separate the compound names into distinct variables:

```{r}
library(tidyr)
who2_long =  who2 |> pivot_longer(
  cols = !(country:year),
  names_to = c("diagnosis", "gender", "age"),
  names_sep = "_",
  values_to = "count"
)
who2_long |> head()
```

### Key take‑aways

* Strings are enclosed in quotes; use double quotes by default and escape special characters like backslash (`\\`), newline (`\n`) or tab (`\t`).  `str_length()` and `str_sub()` let you measure and manipulate substrings.
* Concatenate strings with `str_c()` and build templated strings with `str_glue()`.  Use `coalesce()` to handle missing values and `str_flatten()` to collapse many strings into one.
* Regular expressions describe patterns.  Learn the roles of metacharacters (`.`, `?`, `*`, `+`), character classes (`[...]`) and alternation (`|`), and how to anchor patterns at the start (`^`) or end (`$`) of a string.
* Use `str_detect()` to filter strings, `str_count()` to count occurrences, `str_extract()` to pull out matches, and `str_replace()`/`str_replace_all()` to perform substitutions.  These functions are vectorised and integrate well with dplyr.
* Tidy multiple variables encoded in one string by combining `stringr` with tidyr’s `separate()` and `pivot_longer()`.  This prepares your data for converting character vectors to factors in the next session.
