# Session 9: Variable types II: strings and regular expressions

## Objectives

- **Create and manipulate strings.** Recall that strings are sequences of characters defined by quotes. In R you can use single `'` or double `"` quotes to create strings; the R4DS text recommends double quotes by default, switching to single quotes only when the string itself contains double quotes, and raw strings when a string is otherwise full of backslashes. You'll learn to escape special characters, measure length with `str_length()`, and extract or replace substrings with `str_sub()`.
- **Combine and format strings.** Concatenate strings with `str_c()` (stringr's equivalent to `paste0()`, but stricter about missing values), and build templated strings with `str_glue()`. Learn to collapse a vector of strings into one with `str_flatten()`.
- **Understand regular expressions.** A regular expression (regex) is a concise language for describing patterns within strings. Learn the difference between literal characters and metacharacters, how quantifiers (`?`, `*`, `+`, `{n,m}`) control repetition, how character classes (`[...]`) and their shortcuts (`\d`, `\s`, `\w`) work, and how alternation (`|`) and anchors (`^`, `$`, `\b`) narrow a match.
- **Match, extract, and replace patterns.** Use `str_detect()`, `str_subset()`, and `str_which()` to test or filter strings, `str_count()` to count matches, `str_extract()`/`str_extract_all()`/`str_match()` to pull out matches (including captured groups), and `str_replace()`/`str_remove()` (and their `_all()` variants) to edit strings in place.
- **Use capturing groups and literal matching.** Recognize how parentheses create capturing groups you can reuse in a replacement, and know when to reach for `fixed()` instead of treating a search string as a regex at all.
- **Tidy multiple variables encoded in one string.** Recognize when a single string stores several pieces of information and use `stringr` plus tidyr's `separate_wider_delim()` to split it into proper columns.

## Notes

```{r}
#| message: false
#| warning: false
library(tidyverse)
```

### Creating strings: quotes, escapes, and raw strings

Strings can be created with single or double quotes. Double quotes are preferred by convention; use single quotes only when the string itself contains double quotes, to avoid escaping them. Special characters (backslash, newline, tab) require escaping with `\\`, `\n`, or `\t`.

```{r}
quote_example <- "He said, 'R is great!'"
quote_example

path <- "C:\\Users\\Data\\mydata\n"
path
```

When a string needs many literal backslashes, such as a Windows path or a regex pattern you don't want to double-escape, a **raw string** written as `r"(...)"` treats everything inside as literal text, with no escaping at all:

```{r}
raw_path <- r"(C:\Users\Data\mydata)"
raw_path
```

The length of a string (its number of characters) is computed with `str_length()`, which counts human-visible characters rather than bytes, so it handles multi-byte characters like emoji sensibly:

```{r}
str_length(c("", "abc", "😊"))
```

`str_sub()` extracts or replaces substrings by position. Positive indices count from the start and negative indices count from the end, and the function is vectorized:

```{r}
fruit <- c("banana", "apple", "pear")
str_sub(fruit, 1, 3)      # first three letters
str_sub(fruit, -3, -1)    # last three letters
str_sub(fruit, 1, 1) <- str_to_upper(str_sub(fruit, 1, 1))  # capitalize first letter
fruit
```

### Combining and formatting strings

`str_c()` concatenates strings and returns a character vector; it behaves like `paste0()` but integrates more predictably with dplyr and, unlike `paste0()`, does not quietly turn a missing value into the text `"NA"` (see Example 9.2). A `sep` argument inserts a separator between pieces, and `collapse` further collapses the whole result into a single string.

```{r}
species <- c("Adelie", "Chinstrap", "Gentoo")
n <- c(3, 2, 1)
str_c(n, species, sep = " ", collapse = "; ")
```

Inside a `mutate()`, if any input to `str_c()` is `NA`, the whole result is `NA`; wrap a column in `coalesce(column, "")` first if you would rather treat a missing value as an empty string.

```{r}
df <- tibble(first = c("John", NA, "Jenny"), last = c("Smith", "Nguyen", NA))
df |> mutate(full = str_c(coalesce(first, ""), coalesce(last, ""), sep = " "))
```

`str_glue()` builds templated strings, inserting the value of any R expression written inside `{}`:

```{r}
df |> mutate(message = str_glue("Hello {first} {last}!"))
```

To combine many strings into one, `str_flatten()` collapses a character vector with a given delimiter, which reads more clearly than building the same string up manually:

```{r}
str_flatten(c("a", "b", "c"), collapse = ", ")
```

### Regular expressions: pattern basics

A regular expression describes a *set* of strings rather than one specific string. Most letters and digits match themselves literally; a handful of characters are metacharacters with special meaning:

- `.` matches any single character except a newline.
- Quantifiers control repetition: `?` means zero or one, `*` means zero or more, `+` means one or more, and `{n,m}` means between `n` and `m` repetitions (either bound can be omitted).
- Character classes like `[aeiou]` match any one of the enclosed characters, and `[^0-9]` (a caret at the start) matches any character *not* in the class. `\d`, `\s`, and `\w` are shortcuts for a digit, whitespace, and a "word" character (letter, digit, or underscore), respectively.
- Alternation `|` matches one of several alternatives, so `apple|pear` matches either word.
- Anchors pin a match to a position rather than to any characters: `^` matches the start of a string, `$` matches the end, and `\b` matches a word boundary, the (zero-width) transition between a word character and a non-word character (see Example 9.4).

### Detecting, counting, and extracting patterns

`str_detect()` tests whether each string matches a pattern and returns a logical vector, which pairs naturally with `filter()`. `str_subset()` returns the matching strings themselves rather than a logical vector, and `str_which()` returns their positions. `str_count()` counts how many times a pattern matches within each string. `str_extract()` returns the first match (`str_extract_all()` returns every match, as a list), and `str_replace()`/`str_remove()` replace or delete the first match (`str_replace_all()`/`str_remove_all()` handle every match).

```{r}
#| message: false
library(nycflights13)

set.seed(42)
flights_small <- flights |>
  select(carrier, flight) |>
  slice_sample(n = 5) |>
  mutate(flight_id = str_c(carrier, flight, sep = "-"))

flights_small |> mutate(is_three_digit = str_detect(flight_id, "-\\d{3}$"))
flights_small |> mutate(number_only = str_extract(flight_id, "\\d+"))
```

`\d` matches any digit, so `\d+` extracts one or more consecutive digits; the backslash itself has to be doubled (`\\d`) so that R's own string-escaping rules pass a single backslash through to the regex engine.

```{r}
messy <- c("(202) 555-0198", "+1-303-555-1212")
str_replace_all(messy, "[^0-9]", "")
```

### Capturing groups and backreferences

Wrapping part of a pattern in parentheses creates a **capturing group**, which does two things at once: it groups that piece of the pattern together (useful with quantifiers and alternation), and it remembers exactly what matched so you can reuse it. `str_match()` returns the whole match together with the text captured by each group, as a matrix:

```{r}
dates <- c("2026-01-15", "2026-02-20")
str_match(dates, "(\\d{4})-(\\d{2})-(\\d{2})")
```

Inside a replacement string, `\1`, `\2`, and so on refer back to whatever each group captured, which makes it possible to rearrange a match rather than just deleting or replacing it outright:

```{r}
str_replace(dates, "(\\d{4})-(\\d{2})-(\\d{2})", "\\2/\\3/\\1")
```

### Literal matching with `fixed()`

Sometimes the string you are searching for is not a pattern at all, just literal text that happens to contain regex metacharacters, such as a period, parenthesis, or dollar sign. Wrapping the search string in `fixed()` tells stringr to match it character for character, with no regex interpretation (see Example 9.3 for what happens if you forget to).

```{r}
str_detect(c("cat.txt", "cat"), fixed("."))
```

### Splitting a column into several

When you know the delimiter, tidyr's `separate_wider_delim()` splits one column into several. Suppose a column stores names as `"LAST, First Middle"`:

```{r}
people <- c("SMITH, John A.", "O'NEILL, Anne", "Lee, Chen")

tibble(full = people) |>
  separate_wider_delim(full, delim = ",", names = c("last", "rest")) |>
  mutate(rest = str_trim(rest))
```

`separate_wider_delim()` only consumes the literal delimiter itself, so the piece after a `", "` delimiter keeps its leading space; `str_trim()` cleans up any leading or trailing whitespace left behind by a split like this.

### Working with messy strings in tibbles

Character columns often encode multiple variables in one field. The built-in `who2` dataset has column names like `sp_m_014`, packing together a diagnosis method, gender, and age group; combining `pivot_longer()` with a separator unpacks them into proper variables in one step, exactly as in Session 7.

```{r}
who2_long <- who2 |>
  pivot_longer(
    cols = !(country:year),
    names_to = c("diagnosis", "gender", "age"),
    names_sep = "_",
    values_to = "count"
  )

who2_long |> head()
```

### Fringe cases and common pitfalls

::: {.example title="Example 9.1"}
**Quantifiers are greedy by default, and it shows.**

```{r}
html <- "<a><b>"
str_extract(html, "<.+>")     # greedy: matches as much as possible
str_extract(html, "<.+?>")    # lazy: matches as little as possible
```

`+` (like `*` and `{n,m}`) grabs as much text as it possibly can while still letting the overall pattern match, so `<.+>` starts at the first `<` and only stops at the *last* `>` in the string, swallowing `<a><b>` whole instead of stopping at the first `>`. Adding `?` right after a quantifier (`+?`, `*?`, `??`) makes it lazy instead, matching as little as possible, which is almost always what you actually want when extracting something bounded by delimiters that might repeat.
:::

::: {.example title="Example 9.2"}
**`str_c()` and `paste0()` treat a missing value very differently.**

```{r}
x <- c("a", NA, "c")
str_c("val: ", x)     # NA stays a genuine NA
paste0("val: ", x)    # NA becomes the literal text "val: NA"
```

`str_c()` propagates `NA` the way arithmetic does: if any input piece is missing, the whole result for that element is missing too, which is almost always the behavior you want, since it keeps `is.na()` meaningful on the result. `paste0()` instead coerces `NA` to the three-character string `"NA"` and glues it in like any other text, which can silently turn a genuinely missing value into a string that *looks* like real data (and will not be caught by a later `is.na()` check). If you're translating old base R code that leans on `paste()`/`paste0()` into the tidyverse, this is one of the easiest behavior changes to miss.
:::

::: {.example title="Example 9.3"}
**Forgetting that a search string is a regex changes what "matches" means.**

```{r}
words <- c("cat.txt", "cat", "hat")
str_detect(words, ".")              # "." means "any character," not a literal period
str_detect(words, fixed("."))       # fixed() treats "." as a literal period
```

Every one of these strings contains at least one character, so the unquoted `.` pattern matches all three, which is almost certainly not what someone searching for filenames with a literal dot in them intended. The fix is either `fixed(".")`, which turns off regex interpretation entirely, or escaping the metacharacter directly in the pattern (`"\\."`). This mistake is easy to make with any of the regex metacharacters (`.`, `(`, `)`, `$`, `^`, `+`, `*`, `[`, `]`), not just the period.
:::

::: {.example title="Example 9.4"}
**Without a word boundary, a pattern matches inside unrelated words too.**

```{r}
text <- c("cat", "category", "concatenate", "the cat sat")
str_detect(text, "cat")            # matches inside "category" and "concatenate" too
str_detect(text, "\\bcat\\b")      # only matches "cat" as a whole word
```

The plain pattern `"cat"` matches *any* occurrence of those three letters in sequence, including the middle of "con**cat**enate," which has nothing to do with cats. `\b` matches the zero-width boundary between a word character and a non-word character (or the start/end of the string), so `\bcat\b` only matches "cat" when it is not glued to other letters on either side. Whenever you want to match a whole word rather than a substring, word boundaries are the fix.
:::

### Recap

| **Term** | **Definition** |
|-------------------|-------------------------------------------------------------|
| **Raw string (`r"(...)"`)** | A string literal where nothing is escaped, useful for paths and regexes full of backslashes. |
| **`str_length()` / `str_sub()`** | Measure a string's character length, or extract/replace a substring by position. |
| **`str_c()` vs. `paste0()`** | Both concatenate strings, but `str_c()` propagates `NA` while `paste0()` converts it to the text `"NA"`. |
| **Character class shortcuts** | `\d` (digit), `\s` (whitespace), `\w` (word character: letter, digit, or underscore). |
| **Quantifier** | `?`, `*`, `+`, or `{n,m}`, controlling how many times the preceding piece may repeat; greedy by default, lazy with a trailing `?`. |
| **Word boundary (`\b`)** | A zero-width anchor matching the edge of a word, preventing a pattern from matching inside a longer word. |
| **Capturing group (`(...)`)** | Groups part of a pattern and remembers what it matched, for reuse via `\1`, `\2`, and so on, or extraction with `str_match()`. |
| **`fixed()`** | Tells a stringr function to treat its pattern as literal text rather than a regex. |
| **`separate_wider_delim()`** | Splits one column into several at a literal delimiter. |

### Check your understanding

::: {.callout-note collapse="false" title="Problems"}
1. What is the difference between `str_extract(x, "<.+>")` and `str_extract(x, "<.+?>")` on the string `"<a><b>"`? Which one would you want if you were trying to extract just the first tag?
2. A dataset has a column where missing values were introduced with `str_c()`. Explain why `is.na()` still works correctly on that column, and why the same code using `paste0()` instead might not have.
3. You want to find every row where a `notes` column contains a literal question mark. `str_detect(notes, "?")` returns strange results (or an error). What went wrong, and what are two ways to fix it?
4. Explain what `\bcat\b` matches that plain `cat` does not, and give an example string where the two patterns would disagree.
5. In `str_replace(dates, "(\\d{4})-(\\d{2})-(\\d{2})", "\\2/\\3/\\1")`, what do `\\1`, `\\2`, and `\\3` refer to?
:::

::: {.callout-tip collapse="true" title="Solutions"}
1. `"<.+>"` is greedy and matches the entire string `"<a><b>"`, since `+` grabs as much as possible while still allowing the pattern to match, stopping only at the last `>`. `"<.+?>"` is lazy and matches just `"<a>"`, stopping at the first `>` it can. To extract just the first tag, use the lazy version.

2. `str_c()` propagates missingness: if any piece being concatenated is `NA`, the whole result is `NA`, so `is.na()` still correctly identifies those rows afterward. `paste0()` instead converts a missing value into the literal three-character string `"NA"`, which is not a missing value at all as far as R is concerned, so `is.na()` would return `FALSE` for those rows even though the data started out missing.

3. `?` is a regex metacharacter (a quantifier meaning "zero or one of the preceding thing"), not a literal question mark, so a bare `"?"` pattern is not valid on its own without something for it to quantify (or is silently interpreted as an empty, always-matching pattern, depending on context). Fix it with `str_detect(notes, fixed("?"))`, which turns off regex interpretation entirely, or by escaping it directly: `str_detect(notes, "\\?")`.

4. `\b` anchors to a word boundary, so `\bcat\b` only matches "cat" as a complete, standalone word, not as part of a longer word. In `"concatenate"`, plain `cat` matches the "cat" hidden inside "con**cat**enate," while `\bcat\b` does not match anywhere in that string at all, since "cat" there is not bounded by non-word characters on both sides.

5. `\1` refers to whatever the first parenthesized group captured (the four-digit year), `\2` refers to the second group (the two-digit month), and `\3` refers to the third group (the two-digit day). The replacement string rearranges them into `month/day/year` order using those backreferences instead of the original matched text.
:::
