# Session 13: Relational data

## Objectives

- **Recognize relational data and keys.** A relational dataset organizes information across multiple tables; each table has a primary key that uniquely identifies each row, and a foreign key links a row in one table to a row in another. Learn to identify keys, verify their uniqueness, and create a surrogate key with `row_number()` when no natural key exists.
- **Use mutating joins to combine tables.** Mutating joins (`left_join()`, `inner_join()`, `right_join()`, `full_join()`) add new columns from a matching table by matching rows on keys. Understand how a left join keeps every row of the left table, and how duplicates arise in a many-to-many join.
- **Specify join keys explicitly with `join_by()`.** The default join behavior uses every common variable name as the key, which is convenient right up until two tables happen to share a column name that means two different things. Learn to specify keys manually, including keys with different names on each side.
- **Use filtering joins to filter rows.** Filtering joins (`semi_join()`, `anti_join()`) keep or drop rows in one table based on whether a match exists in another; unlike mutating joins, they never add columns and never duplicate rows.
- **Avoid pitfalls with joins.** Recognize issues such as missing matches, duplicate keys, many-to-many relationships that inflate row counts, and the surprising way a `NA` in a join key is treated.

## Notes

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(nycflights13)
```

### What is relational data?

Data rarely lives in a single flat table. When information naturally splits into multiple tables, you have relational data. The `nycflights13` package is a good example: separate tables describe individual flights, airlines, planes, airports, and weather. To connect them, each table has a **primary key**, a variable (or set of variables) that uniquely identifies each row, and other tables contain **foreign keys** that reference those primary keys. Keys let you work with smaller, more coherent tables instead of duplicating the same information across many columns.

To reliably join tables, check that a candidate key is actually unique within its table, using `count()` and `filter(n > 1)` to surface any duplicates. When no single column is unique on its own, a **compound key** made of several columns together, such as airport and hour for weather observations, can serve the same purpose. If a table genuinely has no natural key at all, `row_number()` creates a simple **surrogate key**, useful mainly for referring to a specific row unambiguously (in a bug report, for instance) rather than for anything relational.

### Mutating joins

A mutating join combines variables from two tables by matching rows on their keys, and the four common mutating joins differ only in which rows survive.

- **Inner join** (`inner_join(x, y)`): keeps only rows where the key appears in both `x` and `y`.
- **Left join** (`left_join(x, y)`): keeps every row of `x`, filling unmatched rows' new columns with `NA`. This is the most common join, since it augments your existing data without silently dropping any of it.
- **Right join** (`right_join(x, y)`): keeps every row of `y`, matching as many rows of `x` as possible.
- **Full join** (`full_join(x, y)`): keeps every row from both tables, filling `NA` wherever there's no match.

```{r}
flights |> left_join(airlines, by = "carrier")
```

Because a mutating join adds columns the same way `mutate()` does, it's easy not to notice new columns arriving if the table already has many; `select()` down to the relevant keys and columns before joining if you want to see exactly what changed.

#### Specifying join keys

By default, dplyr joins on every variable name the two tables have in common, which is convenient exactly until two tables share a column name that doesn't mean the same thing in both places (see Example 13.1). `join_by()` lets you name the key (or keys) explicitly, including an *equi join* where the matching columns have different names on each side:

```{r}
flights |> left_join(airports, join_by(dest == faa))
```

The older syntax `by = "carrier"` or `by = c("playerID" = "id")` still works and does the same thing; `join_by()` is simply the more explicit, more readable modern spelling, and the one you'll need for the non-equi joins mentioned below.

### Filtering joins

A filtering join keeps or drops rows from the first table based solely on whether a match exists in the second, and, unlike a mutating join, never adds a single column and never duplicates a row (see Example 13.3).

- **Semi join** (`semi_join(x, y)`): keeps rows in `x` that have at least one match in `y`. Use it to restrict `x` to cases that also appear somewhere in another table.
- **Anti join** (`anti_join(x, y)`): keeps rows in `x` that have *no* match in `y`. Use it to find observations in `x` that are missing from `y`, exactly the technique previewed for spotting implicit missing values back in Session 12.

```{r}
# flights whose plane has no recorded details in the planes table
flights |> anti_join(planes, by = "tailnum") |> distinct(tailnum) |> head()
```

### Handling missing matches and duplicate keys

Missing matches are inevitable in real data. Left-joining plane characteristics onto flight records leaves some planes' details unknown, filled with `NA`, and it's worth deciding whether that reflects data that genuinely doesn't exist or a foreign key that's simply wrong somewhere upstream.

Duplicate keys are the bigger risk, since a join multiplies rows whenever a key isn't unique on at least one side (see Example 13.2). Check for duplicates before joining, and decide whether you actually want every combination (setting `relationship = "many-to-many"` to say so explicitly) or whether one table needs to be summarized down to one row per key first.

### A glance at more advanced joins

Every join so far has matched rows on exact equality. dplyr's `join_by()` also supports **non-equi joins**: inequality conditions (`join_by(x >= y)`) for matching ranges, `cross_join()` for every possible pairing between two tables, rolling joins (`join_by(closest(x <= y))`) for matching each row to its nearest counterpart rather than an exact one, and overlap joins (`join_by(between(x, y_lower, y_upper))` and similar) for matching a value against a range. These come up most often when matching events to time windows or matching on approximate rather than exact criteria; the four joins covered above handle the large majority of everyday relational work.

### Fringe cases and common pitfalls

::: {.example title="Example 13.1"}
**The default join can silently match on a shared column name that means two different things.**

Both `flights` and `planes` happen to have a column called `year`, but in `flights` it means "year the flight occurred" (every value is `2013`), while in `planes` it means "year the plane was manufactured" (ranging from 1956 to 2013). Joining on every shared column name by default pulls `year` in as part of the key without asking:

```{r}
small_flights <- flights |> select(year, tailnum) |> slice_head(n = 5)

# accidental: joins on tailnum AND year, so it only "matches" a plane
# manufactured in exactly 2013, which is almost never true
small_flights |> left_join(planes, by = c("tailnum", "year")) |> select(tailnum, manufacturer)

# correct: join on tailnum alone
small_flights |> left_join(planes, by = "tailnum") |> select(tailnum, manufacturer)
```

The first join looks like it worked, since it runs without an error or a warning, but nearly every `manufacturer` comes back `NA`, because almost no plane was manufactured the same year as any given 2013 flight. `join_by()` (or an explicit `by =`) removes the ambiguity entirely by naming exactly which column, or columns, should actually be compared. Whenever two tables share a column name, pause and check whether it means the same thing in both places before letting a natural join use it.
:::

::: {.example title="Example 13.2"}
**An unexpected many-to-many relationship inflates your row count, but at least it warns you.**

```{r}
#| warning: true
x <- tibble(id = c(1, 1, 2), val_x = c("a", "b", "c"))
y <- tibble(id = c(1, 1, 2), val_y = c("m", "n", "o"))

left_join(x, y, by = "id")
```

`id = 1` appears twice in both `x` and `y`, so the join produces every combination of the two, four rows for `id = 1` alone, growing a 3-row table into a 5-row result. dplyr detects that this many-to-many situation wasn't declared up front and warns you, which is exactly the kind of warning worth reading rather than dismissing, since a quietly duplicated dataset can throw off every downstream count and average. If the many-to-many relationship is genuinely intended, `relationship = "many-to-many"` silences the warning; if it isn't, the fix is almost always to deduplicate one side of the join first.
:::

::: {.example title="Example 13.3"}
**Filtering joins never add columns, even though it's tempting to expect them to.**

```{r}
names(semi_join(small_flights, planes, by = "tailnum"))
```

The result of `semi_join()` has exactly the same columns `small_flights` started with, nothing from `planes` at all, because a filtering join only ever asks "does a match exist?" and uses the answer to keep or drop rows. If you actually want columns from the second table, you need a mutating join (typically `inner_join()`, if you also want to drop the rows without a match); reach for `semi_join()`/`anti_join()` only when the columns you already have are all you need, and you're just trying to filter based on another table's contents.
:::

::: {.example title="Example 13.4"}
**A `NA` in a join key matches another `NA`, unlike in SQL.**

```{r}
a <- tibble(key = c(1, NA, 3), val_a = c("x", "y", "z"))
b <- tibble(key = c(1, NA, 4), val_b = c("p", "q", "r"))

inner_join(a, b, by = "key")
```

Anyone with a SQL background may expect two `NULL`s to never match each other, since SQL treats `NULL` as "unknown," and two unknowns can't be asserted equal. dplyr's joins work differently: they treat `NA` as an ordinary value for matching purposes, so a row with `key = NA` in `a` does match a row with `key = NA` in `b`. This is rarely what you actually want if `NA` represents "missing" rather than a genuine shared category, so filtering out `NA` keys before joining (`filter(!is.na(key))`) is worth doing explicitly whenever a missing key showing up as a false match would be a real problem.
:::

### Recap

| **Term** | **Definition** |
|-------------------|-------------------------------------------------------------|
| **Primary key** | A variable (or set of variables) that uniquely identifies each row in a table. |
| **Foreign key** | A variable in one table that refers to a primary key in another table. |
| **Surrogate key** | An artificial identifier (typically from `row_number()`) added when no natural key exists. |
| **Mutating join** | Adds columns from a second table by matching keys (`inner_join()`, `left_join()`, `right_join()`, `full_join()`). |
| **`join_by()`** | Specifies join keys explicitly, including keys with different names on each side or non-equality conditions. |
| **Filtering join** | Keeps or drops rows from one table based on whether a match exists in another (`semi_join()`, `anti_join()`); never adds columns or duplicates rows. |
| **Many-to-many relationship** | When a key isn't unique on either side of a join, so matching rows are duplicated for every combination; triggers a warning unless declared with `relationship = "many-to-many"`. |
| **`NA` in a join key** | Matches another `NA` in dplyr's joins, unlike SQL's `NULL`, which never matches another `NULL`. |

### Check your understanding

::: {.callout-note collapse="false" title="Problems"}
1. What is the difference between a primary key and a foreign key? What is a surrogate key, and when would you need one?
2. Two tables both have a column called `id`, but in one table it identifies a customer and in the other it identifies an order. What would happen if you joined them with the default `by` behavior, and how would `join_by()` help?
3. A `left_join()` you expected to simply add a couple of columns instead produces a table with far more rows than you started with, along with a warning. What almost certainly happened?
4. Explain why `semi_join(orders, customers, by = "customer_id")` cannot be used to add a customer's name to the orders table, even though both tables share a `customer_id` column.
5. You join two tables on a key that can be `NA` in both tables, expecting rows with an unknown key to simply be excluded from the match. What actually happens in dplyr, and how would you get the behavior you expected?
:::

::: {.callout-tip collapse="true" title="Solutions"}
1. A primary key uniquely identifies each row within its own table; a foreign key is a column in a different table that refers back to that primary key, linking the two tables together. A surrogate key is an artificial identifier, typically created with `row_number()`, added when a table has no natural column (or combination of columns) that's already unique; it's mainly useful for referring to a specific row unambiguously, not for representing any real-world relationship.

2. By default, the join would use `id` as the key, silently comparing a customer identifier to an order identifier as though they meant the same thing, which would produce meaningless or empty matches without any error. `join_by()` (or an explicit `by =`) lets you name the actual keys you want compared, or simply avoid using `id` as the join key at all if the two columns aren't meant to be compared.

3. The join key almost certainly isn't unique on at least one side, creating a many-to-many relationship: every duplicate key on one side gets matched against every duplicate key on the other, multiplying rows far beyond a simple one-to-one or one-to-many join. Checking each table's key with `count()` and `filter(n > 1)` before joining would reveal the duplicates.

4. `semi_join()` is a filtering join: it only ever keeps or drops rows from the first table based on whether a match exists in the second, and never brings in any columns from the second table at all. To add a customer's name to the orders table, you need a mutating join, such as `left_join(orders, customers, by = "customer_id")`.

5. dplyr treats `NA` as an ordinary matchable value in a join key, so rows with `NA` in both tables' key columns will match each other, the opposite of SQL's `NULL`, which never matches another `NULL`. To exclude rows with a missing key from matching, filter them out of one or both tables first, for example with `filter(!is.na(key))`, before joining.
:::
