# Advanced data import I: spreadsheets

## Objectives

- **Recognize why spreadsheets are common and problematic.** Data often arrives as Excel or Google Sheets files. Compared to CSVs, spreadsheets can contain multiple worksheets, extraneous formatting, and mixed data types. Understand why a clean import requires carefully choosing the right sheet, range, and column names.
- **Load Excel spreadsheets into R.** Use the `readxl` package to read `.xls` and `.xlsx` files with `read_xls()`, `read_xlsx()`, and the convenience wrapper `read_excel()`. Specify column names, skip rows, and define custom missing-value strings when importing.
- **Work with multiple worksheets.** List the sheet names in a workbook with `excel_sheets()`, read individual sheets with `read_excel()`, and combine them with `bind_rows()`. Recognize that a sheet's position in the workbook is not a reliable substitute for its name.
- **Read part of a sheet.** Use the `range` argument to select a specific rectangle of cells when a worksheet contains extraneous text above or below the data.
- **Understand type guessing and specify column types.** Excel cells can contain booleans, numbers, datetimes, or strings; readxl guesses column types but you can override them. Recognize how a stored Excel date can surface as a meaningless raw number instead of an actual date.
- **Write Excel files.** Create spreadsheets with `writexl::write_xlsx()` and know its limitations; know when to reach for `openxlsx` instead for more advanced formatting.
- **Load data from Google Sheets.** Use the `googlesheets4` package to read public or shared spreadsheets; `read_sheet()` mirrors `read_excel()`'s interface closely enough that most of what you learn here transfers directly.

## Notes

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(readxl)
```

### Why spreadsheets?

Plain text formats like CSV are easy to version control and read into R, but many collaborators work in spreadsheets instead. A spreadsheet file can hold multiple worksheets, merge cells for visual layout, and mix formatting in with data. That flexibility is convenient for a person skimming the file, and considerably less convenient for code, which needs to know exactly which worksheet holds the data, whether there are header or footer rows to skip, and whether each column actually represents one single variable.

### Reading Excel spreadsheets

`readxl` provides three functions: `read_xls()` for the older `.xls` format, `read_xlsx()` for the newer `.xlsx` format, and `read_excel()`, which detects the file type automatically. All three return a tibble with a consistent interface: the first argument is a path, and the optional `sheet` argument selects a worksheet by name or position.

```{r}
path <- readxl_example("datasets.xlsx")
excel_sheets(path)   # a workbook can hold several unrelated tables

read_excel(path, sheet = "chickwts") |> head()
```

Real spreadsheets are rarely this tidy. `readxl`'s own `deaths.xlsx` example file has explanatory text scattered above and below the actual data table:

```{r}
#| warning: true
deaths_path <- readxl_example("deaths.xlsx")
read_excel(deaths_path)
```

Reading the sheet with no further guidance pulls in the stray text rows as if they were data, and produces a column-name repair warning worth reading rather than dismissing (see Example 14.2). `skip` (to drop leading rows), `col_names` (to supply your own names outright), and `na` (to mark additional strings as missing) clean this kind of import up, the same way they do for `read_csv()`.

### Reading part of a sheet with `range`

When the real data occupies only a rectangle of the sheet, `range` reads exactly that rectangle and nothing else.

```{r}
read_excel(deaths_path, range = "A5:F15")
```

This is the more direct fix for the `deaths.xlsx` example above: rather than guessing at `skip` and hoping nothing extraneous remains below the data, naming the exact cell range sidesteps the stray text entirely, on both ends.

### Working with multiple worksheets

`excel_sheets()` lists every sheet in a workbook. When the same kind of data is split across several sheets, such as one sheet per site or per month, reading each sheet into its own tibble and combining them with `bind_rows()` is the standard pattern:

```{r}
#| eval: false
path <- "data/penguins.xlsx"
sheet_names <- excel_sheets(path)

penguins <- sheet_names |>
  set_names() |>
  map(~ read_excel(path, sheet = .x, na = "NA")) |>
  bind_rows(.id = "island")
```

Passing `.id = "island"` to `bind_rows()` records which sheet each row came from, which matters here since the sheet name (the island) isn't otherwise a column in any individual sheet.

### Data types and type guessing

Unlike a CSV, an Excel cell natively holds a boolean, a number, a datetime, or text, and readxl guesses each column's type from a sample of its cells, the same way `read_csv()` does. Mixed or unusual values can confuse that guess (see Example 14.3), and `col_types` overrides it explicitly when needed.

Internally, Excel stores every date as a plain number, a count of days since a fixed reference date, and only *displays* it as a date because of cell formatting that a plain data import can't see. When readxl successfully recognizes a column as a date, it converts that number back for you automatically; when it doesn't, you're left holding a meaningless-looking integer instead of a date, with no error to flag it (see Example 14.1).

### Writing Excel files

`writexl::write_xlsx()` writes a data frame to a genuine `.xlsx` file with simple formatting, and reading it back confirms the round trip.

```{r}
library(writexl)

bake_sale <- tibble(
  item = factor(c("brownie", "cupcake", "cookie")),
  quantity = c(10, 5, 8)
)

xlsx_path <- tempfile(fileext = ".xlsx")
write_xlsx(bake_sale, path = xlsx_path)
read_excel(xlsx_path)
```

If you need finer control, multiple worksheets in one file, or cell styling, `openxlsx` provides functions for column widths, fonts, and colors that `write_xlsx()` intentionally leaves out in favor of simplicity.

### Google Sheets

Many teams collaborate in Google Sheets rather than local Excel files. The `googlesheets4` package's `read_sheet()` mirrors `read_excel()`'s interface closely: it accepts `col_names`, `skip`, `na`, and `col_types`, and takes a Google Sheets URL or file ID in place of a local path.

```{r}
#| eval: false
library(googlesheets4)

gs4_deauth()   # skip authentication for a publicly shared sheet

sheet_id <- "1V1nPp1tzOuutXFLb3G9Eyxi3qxeEhnOXUzL5_BcCQ0w"
students <- read_sheet(
  sheet_id,
  col_names = c("student_id", "full_name", "favourite_food", "meal_plan", "age"),
  skip = 1,
  na = c("", "N/A"),
  col_types = "dcccc"
)
```

`read_sheet()` needs an active internet connection, and a handful of functions (`sheet_names()`, `sheet_write()`) exist only on the Google Sheets side with no `readxl` equivalent, but most of what you already know about `read_excel()` transfers directly.

### Fringe cases and common pitfalls

::: {.example title="Example 14.1"}
**A date stored in Excel can surface as a meaningless raw number.**

The `clippy.xlsx` example file lays out its data as label-value pairs rather than one column per variable, which defeats readxl's date-guessing entirely, since it never sees a whole column of consistent dates to recognize:

```{r}
clippy_path <- readxl_example("clippy.xlsx")
read_excel(clippy_path)
```

`"39083"` is not a typo or a strange unit; it's the literal number of days Excel has counted since its internal reference date, with the date formatting that would normally display it as a real date stripped away by this layout. Converting it back is a one-line fix once you know to look for it:

```{r}
as.Date(39083, origin = "1899-12-30")
```

`1899-12-30` (not 1900-01-01) is Excel's actual epoch, offset by a long-standing historical bug in Excel's leap-year handling that Microsoft never fixed, for backward compatibility. Whenever a numeric column is suspiciously close to five digits and the surrounding columns are clearly about dates, suspect this before assuming the number means something else entirely.
:::

::: {.example title="Example 14.2"}
**Blank or duplicate header cells get silently renamed, and the warning is easy to miss.**

Reading `deaths.xlsx` without a `range` pulls in a row of genuinely blank header cells above the real data:

```{r}
#| warning: true
read_excel(deaths_path) |> names()
```

readxl can't leave a column with no name at all, so it invents `...2`, `...3`, and so on, and reports having done so in a "New names" message that looks more like routine console output than an actual warning. If you don't look closely, it is easy to proceed with a data frame whose columns are named `...2` through `...6` without ever noticing that the real headers were sitting a few rows further down, exactly where `range = "A5:F15"` (used earlier in this session) finds them correctly.
:::

::: {.example title="Example 14.3"}
**A handful of inconsistent values can collapse an entire column's guessed type to character.**

`readxl`'s `type-me.xlsx` example is built specifically to demonstrate this: a column that's mostly boolean-looking values, with a few outliers mixed in.

```{r}
type_path <- readxl_example("type-me.xlsx")
read_excel(type_path, sheet = "logical_coercion")
```

Values like `TRUE`, `FALSE`, `0`, and `1` all look like they belong in a logical column, but `"cabbage"` and `"F"` (a bare letter, not `FALSE`) do not, so readxl falls back to reading the *entire* column as character rather than guessing logical and silently mangling the outliers. This is the same all-or-nothing behavior you saw with `read_csv()`'s numeric guessing back in Session 7: one inconsistent value is enough to change how every other value in the column gets interpreted, not just its own cell.
:::

::: {.example title="Example 14.4"}
**Reading a sheet by position instead of by name can silently give you the wrong table.**

```{r}
excel_sheets(path)   # recall: "mtcars", "chickwts", "quakes", in that order

read_excel(path, sheet = 1) |> names()     # not chickwts!
read_excel(path, sheet = "chickwts") |> names()
```

`sheet = 1` reads whichever sheet happens to be first in the workbook, which is `mtcars` here, not `chickwts`. Nothing about this errors or warns, because `sheet = 1` is a perfectly valid request; it just isn't the request you meant if you assumed a specific dataset would always be first. Referring to a sheet by its name rather than its position is worth the extra few characters of typing, especially for a workbook someone else maintains and might reorder later without telling you.
:::

### Recap

| **Term** | **Definition** |
|-------------------|-------------------------------------------------------------|
| **`read_excel()`** | Reads an Excel file (`.xls` or `.xlsx`), auto-detecting the format; `sheet` selects a worksheet by name or position. |
| **`excel_sheets()`** | Lists every worksheet name in a workbook, without reading any of the data. |
| **`range`** | Reads only a specific rectangle of cells (for example, `"A5:F15"`), skipping extraneous rows or columns entirely. |
| **Column name repair** | readxl automatically renames blank or duplicate header cells to `...2`, `...3`, and so on, reporting the change in a "New names" message. |
| **Excel date storage** | Internally just a count of days since `1899-12-30` (Excel's epoch); displayed as a date only through cell formatting that a plain import can't see. |
| **`col_types`** | Overrides readxl's guessed type for one or more columns, the same role `col_types` plays in `read_csv()`. |
| **`write_xlsx()`** | Writes a data frame to a simple `.xlsx` file; use `openxlsx` for advanced formatting `writexl` doesn't support. |
| **`read_sheet()`** | `googlesheets4`'s equivalent of `read_excel()`, reading a Google Sheet by URL or ID over the internet. |

### Check your understanding

::: {.callout-note collapse="false" title="Problems"}
1. A workbook has explanatory text above and below the actual data table on its only worksheet. Name two different `read_excel()` arguments that could clean this up, and explain how they differ.
2. A colleague imports a spreadsheet and finds a column full of numbers like `44561` where they expected dates. What almost certainly happened, and how would they fix it?
3. `read_excel()` prints a "New names" message listing `...2` and `...3` after you import a sheet. What triggered this, and what should you do about it?
4. A column that should be logical (`TRUE`/`FALSE`) comes back from `read_excel()` as character instead. What is the most likely explanation, and how would you find the specific values responsible?
5. Why is `read_excel(path, sheet = "sales")` generally safer than `read_excel(path, sheet = 2)`, even if sheet 2 happens to be named "sales" today?
:::

::: {.callout-tip collapse="true" title="Solutions"}
1. `skip` drops a fixed number of rows from the top of the sheet before reading, which works well when the extraneous content is only above the data. `range` instead reads one specific rectangle of cells, handling extraneous content both above and below (or on either side of) the data in a single argument, without needing to know exactly how many rows to skip.

2. The column almost certainly holds genuine Excel dates that readxl did not recognize as dates, so what's showing up is the raw internal representation, a count of days since Excel's epoch. Converting it with `as.Date(x, origin = "1899-12-30")` recovers the actual date.

3. The sheet had one or more blank (or duplicate) header cells in its first row, and readxl cannot leave a column with no name at all, so it invented placeholder names. This is a signal that you probably haven't found the real header row yet; check whether a `range` or `skip` argument would land you on the actual header instead.

4. Some value in that column doesn't look like a boolean at all (a stray word, a letter that isn't exactly `TRUE`/`FALSE`, or similar), so readxl fell back to reading the whole column as character rather than guessing logical and mangling just the one inconsistent cell. Reading the column in and then filtering for values that aren't `"TRUE"`, `"FALSE"`, `"0"`, or `"1"` (as text) would surface the specific offending values.

5. A sheet's position in a workbook can change if someone inserts, deletes, or reorders sheets later, silently pointing `sheet = 2` at a completely different table with no error or warning. A sheet's name is far less likely to change casually, and if it does change, code that refers to it by name fails loudly (with a clear "sheet not found" error) instead of silently reading the wrong data.
:::
