# Session 15: Advanced data import II: databases

## Objectives

- **Motivate working directly with databases.** Large volumes of data often live in databases instead of flat files, and repeatedly exporting CSV snapshots is time-consuming. Learn why connecting directly to a database gives you immediate access to current data and avoids that delay.
- **Explain how database tables differ from tibbles.** Database tables live on disk, can be arbitrarily large, and usually have indexes to speed up queries. Classical databases prioritize writing over analysis, whereas column-oriented systems like duckdb are optimized for analytical queries.
- **Connect to a database with DBI.** The DBI package provides a uniform interface to many database systems. Learn to call `DBI::dbConnect()` with a driver from a package like `RPostgres`, `RMariaDB`, or `duckdb`, supplying host, port, and credentials. We will use duckdb, since it runs in-process and needs no server.
- **Load and inspect tables.** Use `dbWriteTable()` to copy a tibble into a database, `dbListTables()` to see what tables exist, and `dbReadTable()` to read a whole table back into R.
- **Execute SQL queries via DBI.** Write simple `SELECT ... FROM ... WHERE ...` statements and run them with `dbGetQuery()`. Understand the core SQL clauses and how they map to dplyr verbs.
- **Use dbplyr for lazy pipelines.** dbplyr translates dplyr code into SQL. Create a lazy table with `tbl()`, compose pipelines with verbs like `filter()` and `summarize()`, view the generated SQL with `show_query()`, and retrieve results with `collect()`. Recognize that a lazy query and an actual data frame are not interchangeable until you `collect()`.

## Notes

### Why work with databases?

When a colleague hands you data by exporting a CSV, you have to go back and ask for a new export every time your question changes. Connecting directly to a database avoids that entirely: rather than working from a snapshot, you write a query that retrieves exactly the current data you need. A database table looks like a data frame, a collection of named columns, but differs in a few important ways. Data frames live in memory; database tables live on disk and can be arbitrarily larger than your computer's RAM. Most databases maintain indexes that let them locate rows quickly without scanning the whole table. And row-oriented databases are built to record individual transactions quickly, while column-oriented systems like duckdb are built for exactly the kind of broad analytical query this course focuses on.

### Setting up a connection

DBI defines a low-level interface for connecting to many different database systems. You create a connection with `DBI::dbConnect()`, supplying a driver from a DBMS-specific package along with connection details. Connecting to a real PostgreSQL server, for example, looks like this (not run here, since it needs an actual server to connect to):

```{r}
#| eval: false
con <- DBI::dbConnect(
  RPostgres::Postgres(),
  host = "db.example.com",
  port = 5432,
  dbname = "mydb",
  user = "username",
  password = "secret"
)
```

This course uses duckdb instead, an in-process database that lives entirely within R and needs no server or credentials at all, which makes it ideal for learning:

```{r}
#| message: false
library(tidyverse)
library(DBI)
library(duckdb)

con <- dbConnect(duckdb())
```

Passing a `dbdir` argument to `duckdb()` points the database at a file on disk instead, so its contents persist between R sessions rather than disappearing (as an in-memory database does) the moment you disconnect.

### Loading and inspecting tables

A brand-new database is empty, so you need to load data into it before you can query anything. `dbWriteTable()` copies a tibble into the database: the connection first, then a table name, then the data frame itself.

```{r}
dbWriteTable(con, "mpg", ggplot2::mpg)
```

`dbListTables()` lists what's actually in the database, and `dbReadTable()` reads an entire table back into R as an ordinary data frame.

```{r}
dbListTables(con)
dbReadTable(con, "mpg") |> as_tibble()
```

### Running SQL queries

If you already know SQL, `DBI::dbGetQuery()` runs a query string directly and returns the result as a data frame. A query is built from clauses like `SELECT`, `FROM`, `WHERE`, `GROUP BY`, and `ORDER BY`.

```{r}
sql <- "SELECT manufacturer, model, year, hwy
        FROM mpg
        WHERE hwy > 30
        ORDER BY hwy DESC"

dbGetQuery(con, sql) |> head()
```

SQL gives you complete control, but writing it by hand is verbose for anything beyond a simple query. dbplyr lets you write the dplyr code you already know instead.

### Lazy queries with dbplyr

`tbl()` creates a *lazy table*: something that represents a database table and behaves like one for the purpose of writing dplyr code, without actually pulling any data into R yet.

```{r}
mpg_db <- tbl(con, "mpg")
mpg_db
```

You compose a query with ordinary dplyr verbs, and dbplyr silently translates each one into SQL behind the scenes, only actually running anything once you call `collect()`.

```{r}
summary_db <- mpg_db |>
  filter(displ >= 2.0) |>
  group_by(manufacturer) |>
  summarise(avg_hwy = mean(hwy, na.rm = TRUE), n = n()) |>
  arrange(desc(avg_hwy))

summary_db |> show_query()

summary_db |> collect()
```

`show_query()` prints the SQL dbplyr generated, without running it; `collect()` actually sends that SQL to the database, retrieves the result, and hands it back as a genuine tibble. Until you call `collect()`, you're working with a description of a query, not with data (see Example 15.4).

### Basic SQL structure

Even though dbplyr hides most of SQL from you, it helps to recognize its shape. A query is built from, at minimum, these clauses, always in this order: `SELECT` (which columns, and any computed expressions), `FROM` (which table), `WHERE` (which rows), `GROUP BY` (how to collapse rows into groups), and `ORDER BY` (how to sort the result). SQL keywords are case-insensitive, but convention writes them in uppercase to stand out from table and column names.

### Fringe cases and common pitfalls

::: {.example title="Example 15.1"}
**A lazy query's `mean()` quietly matches SQL's behavior, not R's.**

```{r}
#| warning: true
dbWriteTable(con, "nas", tibble(x = c(1, 2, NA, 4)))
nas_db <- tbl(con, "nas")

nas_db |> summarise(avg = mean(x)) |> collect()   # no na.rm at all
mean(c(1, 2, NA, 4))                              # the equivalent plain R calculation
```

In plain R, `mean()` propagates `NA` unless you explicitly add `na.rm = TRUE`. In a dbplyr query, `mean()` gets translated to SQL's `AVG()`, which silently ignores `NULL` values by design, so the lazy query above returns the same answer as `mean(x, na.rm = TRUE)` would, without you ever asking for that behavior. dbplyr does warn about this the first time you hit it in a session, but the warning is easy to miss, and the two functions named `mean()` (R's and SQL's, standing behind it) genuinely disagree about what to do with a missing value. Get in the habit of writing `na.rm = TRUE` explicitly in a lazy query, the same as you would anywhere else, so the behavior doesn't depend on which engine happens to be running the calculation.
:::

::: {.example title="Example 15.2"}
**Filtering on a column you just created can force dbplyr to build a nested subquery.**

```{r}
mpg_db |>
  mutate(kpl = hwy * 0.425) |>
  filter(kpl > 12) |>
  show_query()
```

In dplyr, you can create `kpl` with `mutate()` and immediately `filter()` on it in the very next line, because each step runs in order. SQL's `WHERE` clause, though, is evaluated *before* its `SELECT` clause, so a plain, single-level query has no way to filter on an alias defined in that same `SELECT`. dbplyr works around this automatically by wrapping the first query as a subquery and filtering the *outer* query instead, which is exactly the nested `SELECT ... FROM (SELECT ...)` structure you see above. This is a good reason to actually read `show_query()`'s output occasionally: a pipeline that looks simple in dplyr can compile into something considerably less simple in SQL, for reasons that have nothing to do with your code being wrong.
:::

::: {.example title="Example 15.3"}
**`show_query()` can print syntactically plausible SQL for a function that doesn't actually exist in the database.**

```{r}
my_custom_fun <- function(x) x * 2 + 1

weird <- mpg_db |> mutate(x = my_custom_fun(hwy))
weird |> show_query()
```

```{r}
#| error: true
weird |> collect()
```

dbplyr doesn't know what `my_custom_fun()` is, so it does the only thing it can: it writes the function call into the generated SQL literally, exactly as written, and hopes the database recognizes it. `show_query()` happily shows you that SQL, since generating it doesn't require running anything. Only `collect()` (or printing the lazy table, which collects a preview behind the scenes) actually sends the query to the database, which is the point at which it fails with a real "function does not exist" error. A lazy query that *looks* fine when you inspect its SQL is not the same guarantee as a lazy query that will actually run.
:::

::: {.example title="Example 15.4"}
**A lazy table doesn't know its own row count until you collect it.**

```{r}
lazy <- mpg_db |> filter(hwy > 30)
class(lazy)
nrow(lazy)
```

```{r}
collected <- collect(lazy)
class(collected)
nrow(collected)
```

`nrow()` on a lazy table returns `NA`, not an error and not a real count, because dbplyr has only recorded what query *would* count the matching rows, not actually run it. Counting the rows would mean executing the query, which is exactly the work `collect()` is for. Code that assumes a lazy table behaves identically to a real data frame (checking `nrow()` to decide whether to proceed, for instance) needs a `collect()` first, or an explicit `tally()`/`count()` query, to get an actual number back.
:::

### Recap

| **Term** | **Definition** |
|-------------------|-------------------------------------------------------------|
| **`dbConnect()`** | Opens a connection to a database, given a driver and connection details. |
| **`dbWriteTable()` / `dbReadTable()`** | Copy a tibble into a database table, or read an entire table back into R. |
| **`dbGetQuery()`** | Runs a raw SQL query string and returns the result as a data frame. |
| **`tbl()`** | Creates a lazy table representing a database table, without pulling any data into R. |
| **`show_query()`** | Prints the SQL a lazy dplyr pipeline would run, without running it. |
| **`collect()`** | Actually sends a lazy query to the database and returns the result as a tibble. |
| **`AVG()` vs. `mean()`** | SQL's `AVG()` silently ignores `NULL`s, unlike R's `mean()`, which propagates `NA` unless told `na.rm = TRUE`. |
| **Subquery** | A nested `SELECT` dbplyr generates automatically when SQL's clause order can't otherwise express a dplyr pipeline (such as filtering on a just-created column). |
| **Untranslated function** | A function dbplyr doesn't recognize is passed through into the generated SQL literally, and only fails once the database actually tries to run it. |

### Check your understanding

::: {.callout-note collapse="false" title="Problems"}
1. What is the practical difference between calling `show_query()` and calling `collect()` on the same lazy dplyr pipeline?
2. A teammate computes `tbl(con, "sales") |> summarise(avg_price = mean(price)) |> collect()` on a column with some missing prices, and gets a number back rather than `NA`. Explain why, given that the equivalent plain-R calculation on a vector with a missing value would return `NA`.
3. `mpg_db |> mutate(kpl = hwy * 0.425) |> filter(kpl > 12) |> show_query()` produces a query with a nested `SELECT` inside another `SELECT`, even though the dplyr code has no explicit nesting. Why does dbplyr generate a subquery here?
4. You write `mpg_db |> mutate(z = my_function(hwy))` using a function that doesn't exist in SQL. At what point does this actually fail, and why not sooner?
5. Why does `nrow()` return `NA` on a lazy table instead of the actual number of matching rows?
:::

::: {.callout-tip collapse="true" title="Solutions"}
1. `show_query()` only displays the SQL a pipeline *would* run, without contacting the database at all. `collect()` actually sends that SQL to the database, executes it, and returns the result as a real tibble in R's memory. You can call `show_query()` as many times as you like with no cost; `collect()` is the step that does the actual work.

2. `mean()` inside a dbplyr pipeline gets translated to SQL's `AVG()` function, which silently ignores `NULL` (missing) values as part of its standard behavior, unlike R's own `mean()`, which propagates `NA` unless you add `na.rm = TRUE`. The teammate's query is quietly behaving like `mean(price, na.rm = TRUE)` even though nothing in the code asked for that.

3. SQL evaluates a query's `WHERE` clause before its `SELECT` clause, so a single-level query has no way to filter on `kpl`, an alias that's only defined in the `SELECT`. dbplyr works around this by generating an inner query that computes `kpl`, wrapping it as a subquery, and filtering that outer result instead, since by then `kpl` is an ordinary column of the subquery's output.

4. It fails when the query is actually executed against the database, which happens at `collect()` (or when printing the lazy table, since that collects a preview). `show_query()` alone doesn't fail, because generating the SQL text doesn't require the database to check whether every function mentioned in it actually exists; dbplyr just writes an unrecognized function call into the SQL literally and lets the database be the one to reject it.

5. `nrow()` would need to actually count the matching rows, which means running the query, and a lazy table hasn't been run yet; it's only a description of a query. `NA` signals "unknown until you execute this," rather than dbplyr guessing or silently running the query just to answer `nrow()`. Calling `collect()` first (or running an explicit counting query) gives you a real number.
:::
