# Session 19 – Basic programming II: iteration & functional programming

## Objectives

* **Recognize how iteration differs in R.**  In many programming languages you must explicitly loop over the elements of a vector.  R is different: doubling a numeric vector is as simple as `2 * x` because the operation is *vectorised*.  You’ve already encountered implicit iteration through tools like `facet_wrap()`/`facet_grid()` for faceting, `group_by()` + `summarize()` for group‑wise summaries and the `unnest_*()` functions for list‑columns.  In this session you’ll learn more general tools for explicit iteration.
* **Modify multiple columns with `across()`.**  The `across()` function lets you apply one or more functions to a set of columns in `summarize()` or `mutate()`.  You’ll learn to select columns with `.cols`, supply functions via `.fns`, customize the output names via `.names` and use variants like `if_any()`/`if_all()` for filtering.
* **Read and combine many files with purrr.**  `purrr::map()` iterates over a vector or list and applies a function to each element, returning a list.  You’ll learn a three‑step pattern for importing a directory of files: list the file paths with `list.files()`, map a read function over those paths and then combine the results with `purrr::list_rbind()`.  You’ll see how to extract data encoded in file names using `set_names()` and the `names_to` argument of `list_rbind()`.
* **Handle failures and heterogenous data.**  When iterating over many inputs it’s common for a few to fail.  You’ll use `purrr::possibly()` to wrap a function so that failures return a default (e.g., `NULL`) instead of stopping the whole iteration.  You’ll also learn strategies for dealing with lists of data frames that don’t all have the same structure.
* **Save multiple outputs.**  The same iteration patterns can help when exporting data.  You’ll briefly see how to write multiple data frames to disk or a database and how to generate a series of plots programmatically.

## Notes

### Introduction: implicit and explicit iteration

**Iteration is everywhere.**  In R, many iterative operations are built into the language.  The introduction to the iteration chapter reminds us that doubling a vector uses a single expression (`2 * x`) instead of a loop.  Earlier tools you’ve used—faceting, group‑wise summarization and unnesting list‑columns—perform the same action for many subsets or elements.  These conveniences hide the loop, but sometimes you need more control.  Functional programming tools take functions as inputs and allow you to express iteration concisely.  We will focus on three common tasks: **modifying multiple columns**, **reading multiple files** and **saving multiple outputs**.

### Modifying multiple columns with `across()`

Suppose you have a tibble with several numeric variables and you want to compute a summary for each.  You could write the same function call repeatedly—violating the rule to never copy‑and‑paste more than twice.  Instead, use **`across()`** inside `summarize()` or `mutate()` to iterate over columns.  `across()` has three key arguments:

1. **`.cols`** selects which columns to operate on.  It uses the same tidy‑select syntax as `select()`, so you can use helpers such as `starts_with()`, `ends_with()`, `everything()` and `where()`.  For example, `where(is.numeric)` selects all numeric columns and `!where(is.numeric)` selects everything else.
2. **`.fns`** supplies the function(s) to apply.  For a single function you pass its bare name—without `()`—and `across()` will call it for each column.  To provide additional arguments, wrap the function in an anonymous function `\(x) ...`.  You can supply multiple functions by passing a named list; the names determine the suffixes used in the output.
3. **`.names`** controls the names of the output columns.  By default, the name of each output is a glue specification `{.col}_{.fn}`; you can provide your own specification, e.g., `"{.fn}_{.col}"` to put the function name first.  When using `across()` inside `mutate()`, remember that without `.names` the results will overwrite existing columns; specify new names to preserve the originals.

Here are some common patterns using `across()`:

```{r}
#| message: false
library(tidyverse)

# summarise each numeric column and count rows
iris |> 
  summarize(
    across(where(is.numeric), mean),
    n = n()
  )

# compute mean and standard deviation for numeric columns
iris |>
  summarize(
    across(where(is.numeric), list(mean = ~mean(.x, na.rm = TRUE), 
                                  sd   = ~sd(.x, na.rm = TRUE)),
           .names = "{.fn}_{.col}")
  )

# replace missing values with zero but keep originals
df_miss = tibble(a = c(1, NA, 3), b = c(4, 5, NA))
df_miss |> 
  mutate(across(a:b, ~coalesce(.x, 0), .names = "{.col}_na0"))
```

When filtering, `if_any()` and `if_all()` are handy variants: `filter(if_any(where(is.na)))` keeps rows with at least one missing value, whereas `filter(if_all(where(is.na)))` keeps rows where *all* selected columns are missing.

The fact that `.cols` uses tidy‑select means you can write helper functions that operate on a user‑supplied set of columns.  For example, the `summarize_means()` helper from the book computes the mean of selected columns and drops grouping.  Use embracing (`{{ }}`) to allow column arguments inside your functions.

### Reading multiple files with `purrr::map()`

The second iteration task is combining data stored in many files.  A reliable pattern has three steps:

1. **List files.**  Use `list.files()` to get a character vector of file paths.  Supply the directory with `path`, restrict the file types with a regular‑expression `pattern` like `"[.]csv$"` and set `full.names = TRUE` so the results include the directory.
2. **Read each file.**  Use `purrr::map()` to apply a reading function (e.g., `read_csv()` or `read_excel()`) to each path.  `map(x, f)` is shorthand for calling `f()` on each element of `x` and returning a list.  If you need to pass additional arguments, supply an anonymous function `\(path) read_csv(path, skip = 2)`.
3. **Combine the results.**  Once you have a list of data frames, use `purrr::list_rbind()` to stack them into a single tibble.  When file names contain data, call `set_names()` to give the list elements informative names and use the `names_to` argument of `list_rbind()` to store those names as a column.  You can extract numbers from file names with `readr::parse_number()`.

Here is a minimal example that reads multiple CSVs from a directory, records the year from the file name and combines them:

```{r}
#| eval: false
paths = list.files("data/gapminder", pattern = "[.]csv$", full.names = TRUE)

gapminder = paths |> 
  set_names(basename) |>                           # name each element by file name
  map(readr::read_csv) |>                          # read each file into a tibble
  list_rbind(names_to = "file_name") |>           # combine and create file_name column
  mutate(year = readr::parse_number(file_name))    # extract year from the name
```

When the data frames are heterogeneous (different columns or types), inspect them before binding.  A helper like `df_types()` that records column names, types and missing values can be mapped over the list and the results compared.

### Handling failures

`purrr::map()` stops on the first error.  If any file fails to read, nothing is returned.  To avoid losing all progress you can wrap the reading function with `purrr::possibly()`, which returns a specified default (e.g., `NULL`) instead of erroring.  For example:

```{r}
#| eval: false
safe_read = possibly(readr::read_csv, otherwise = NULL)
files  = paths |> map(safe_read)
data   = files |> list_rbind()    # list_rbind() silently drops NULLs

failed = map_lgl(files, is.null)
paths[failed]  # investigate these files manually
```

This pattern lets you process all the valid files and identify which ones need special handling.

### Saving multiple outputs

The final task is the reverse of reading: you often need to save multiple data frames or plots.  The iteration tools you’ve learned apply here too.  For example, to save several datasets to disk you could use `purrr::iwalk()` (or `walk2()`) over a list of tibbles and corresponding file names:

```{r}
#| eval: false
datasets = list(penguins = palmerpenguins::penguins, 
                 storms   = dplyr::storms)

paths = paste0(names(datasets), ".csv")
walk2(datasets, paths, readr::write_csv)
```

Similarly, to save multiple plots you can iterate over a list of ggplot objects and call `ggsave()` for each, constructing the file name from the plot’s position or a meaningful label.  When writing many outputs to a database, you can create a template table and then iterate over your data frames inserting them one by one.

### Key take‑aways

* **Vectorised operations hide loops.**  R often does the iteration for you, but when you need more control, functional programming tools help you express iteration clearly.
* **`across()` loops over columns.**  Select variables with tidy‑select in `.cols`, supply one or more functions via `.fns` (bare names or anonymous functions), and control output names with `.names`.  Use `where()` helpers to select columns by type and `if_any()`/`if_all()` to filter across multiple columns.
* **Break up complex summaries.**  When multiple functions are applied, name them to get informative suffixes and, if needed, customise the naming pattern.
* **Three‑step file import.**  Use `list.files()` to collect file names, `purrr::map()` to read each file and `list_rbind()` to combine them.  Use `set_names()` and the `names_to` argument to capture metadata stored in file names.
* **Use `possibly()` to handle errors.**  Wrap functions with `purrr::possibly()` so that failures return a default, allowing you to process good inputs and diagnose problems later.
* **Iteration patterns are symmetrical.**  The same tools you use to read multiple inputs—`map()`, `walk()`, naming and combining—also help you write multiple outputs, from CSV files to plots.

<!-- ## In‑class activity -->

<!-- In today’s hands‑on activity you will practise functional programming with the **palmerpenguins::penguins** dataset.  The dataset includes measurements for three species of penguins (`Adelie`, `Gentoo`, `Chinstrap`) and provides a clean, tabular example that is *not used in the R for Data Science book*.  The goal is to apply `across()` to summarise and transform multiple variables, iterate over groups with `map()`, handle potential read failures with `possibly()`, and save multiple outputs.  Complete the following steps in an R script or notebook: -->

<!-- 1. **Explore the data (5 min).**  Load the tidyverse and palmerpenguins packages.  Use `glimpse(penguins)` to see the structure of the dataset, and identify which variables are numeric and which are factors.  Use `summary()` to check for missing values. -->

<!-- 2. **Summarise numeric variables by species (10 min).**  Use `group_by(species)` and `summarize()` with `across()` to compute the mean and standard deviation of every numeric column (e.g., `bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`, `body_mass_g`).  Supply a named list of functions so that your output columns are named `mean_bill_length_mm`, `sd_bill_length_mm`, etc.  Remember to set `na.rm = TRUE`. -->

<!-- 3. **Create standardized variables (10 min).**  Use `mutate()` with `across(where(is.numeric), ~ (.x - mean(.x, na.rm = TRUE)) / sd(.x, na.rm = TRUE), .names = "z_{.col}")` to add Z‑score versions of each numeric variable to the dataset.  Compare the distributions of the standardized variables across species with a boxplot. -->

<!-- 4. **Split and save the data (10 min).**  Split the standardized penguin data by species using `split(penguins, penguins$species)`.  Use `iwalk()` (or `walk2()`) to iterate over the list and write each tibble to a CSV file named after the species, e.g., `"adelie.csv"`, `"gentoo.csv"` and `"chinstrap.csv"`.  All files should be saved in a new folder called `species_csv`. -->

<!-- 5. **Read and combine with error handling (15 min).**  Use `list.files()` with a pattern that matches only `.csv` files in `species_csv` and `full.names = TRUE` to create a vector of file paths.  Create a safe reading function `safe_read = purrr::possibly(readr::read_csv, otherwise = NULL)` and use `map()` to read each file.  Combine the non‑NULL results with `list_rbind(names_to = "species")` to reconstruct a single tibble.  Verify that the combined data matches your original standardized dataset.  If any files fail to read (e.g., you could intentionally rename one file to simulate an error), identify which ones by checking which list elements are `NULL`. -->

<!-- 6. **Generate and save multiple plots (10 min).**  For each species, create a scatterplot of bill length vs. flipper length using `ggplot()` and store the plots in a named list.  Then use `iwalk()` to iterate over the plots and save each to a PNG file in a `plots` folder.  Construct the file names using the species names (e.g., `"scatter-adelie.png"`).  Optionally, customise the titles and colours for each plot. -->

<!-- This activity will reinforce your understanding of `across()`, `map()`, safe iteration with `possibly()` and iteratively saving outputs. -->

<!-- --- -->

<!-- **Suggested reading for Session 20:**  Before the next class on reproducible research and communication, read two chapters from *R for Data Science* (2nd edition).  **Chapter 28 “Quarto”** introduces Quarto as a unified authoring framework that combines code, its results and your prose into a single, reproducible document【992205213102681†L125-L143】.  The chapter explains that Quarto documents (`.qmd` files) are plain text files with a YAML header, code chunks and narrative text, and that Quarto supports many output formats, such as HTML, PDF and Word【992205213102681†L165-L201】.  You’ll learn that Quarto is designed for three purposes: communicating results to decision‑makers, collaborating with other data scientists, and serving as a lab notebook【992205213102681†L125-L142】.  The chapter also notes that Quarto unifies the R Markdown ecosystem and requires the command‑line interface (CLI) but is integrated into RStudio【992205213102681†L149-L160】. -->

<!-- **Chapter 11 “Communication”** focuses on turning exploratory graphics into expository graphics that effectively communicate your insights.  The authors emphasize that your audience may lack your background knowledge, so you need to invest effort in making plots self‑explanatory【778553707203566†L124-L135】.  Start by adding good labels with `labs()`; titles should summarise the main finding rather than merely describing the plot【778553707203566†L183-L185】.  The chapter also explains how to add subtitles and captions to provide context and data sources【778553707203566†L187-L191】.  Reading these sections will prepare you to use Quarto to document your analysis and to design graphics that communicate clearly. -->
