# Session 20 – Reproducible research & communication

## Objectives

* **Appreciate the value of reproducible research.**  Understand why combining your code, its results and your prose in a single document improves transparency, facilitates collaboration and serves as a lab notebook.  Recognize that Quarto integrates the lessons of the R Markdown ecosystem and is designed for analysts, collaborators and decision‑makers.
* **Learn the basics of Quarto documents.**  Create and render a Quarto (`.qmd`) file with a YAML header and interleaved code and narrative.  Understand how to specify global settings in the YAML, supply metadata like title and date, and set the default output format.
* **Use chunk options effectively.**  Control evaluation, display of code, warnings and messages with chunk options (`eval:`, `include:`, `echo:`, `message:` and `warning:`).  Set global chunk options and learn how inline code can report dynamic values inside your prose.
* **Generate multiple output formats.**  Configure Quarto to render the same document to HTML, PDF and Word and learn how to customize options like the table of contents or code‑folding in the YAML.  Understand the difference between persistent settings in the YAML header and transient settings supplied to `quarto_render()`.
* **Communicate your findings clearly.**  Apply principles from the communication chapter to turn exploratory graphics into expository graphics.  Write titles that summarize the main finding rather than simply describing the plot; add subtitles and captions; label axes and legends with units and descriptive names; and use themes, annotations and layout to guide your audience.

## Notes

### Why reproducible research?

In an exploratory notebook it’s easy to lose track of which code produced which results.  Reproducible research addresses this by knitting your analysis and narrative together so that anyone—including future you—can rerun the analysis and obtain the same outputs.  This approach, sometimes called *literate programming*, intersperses code with human‑readable explanations.  In the tidyverse, the **Quarto** system is the modern successor to R Markdown.  A Quarto document is a plain text file with a `.qmd` extension that contains three elements:

1. A **YAML header** delimited by triple dashes (`---`) at the top of the file.  Here you set metadata such as the document title, author, date and default output format.  You can also set default execution options (e.g., hide code globally) in this header.
2. **Narrative text** written in Markdown.  Use headings to structure your document, lists for bullets and paragraphs for explanations.  You can insert inline R code by enclosing it in backticks preceded by `r` (e.g., ``r nrow(data)``) to dynamically insert values into your prose.
3. **Code chunks** that contain R code.  Each chunk begins with triple backticks and a language engine (e.g., ```{r}```) and ends with triple backticks.  You can assign a label and specify options immediately after the opening fences.  When you render the document, Quarto executes each chunk in order and inserts the output.

This structure ensures that anyone with the data and the `.qmd` file can reproduce your results.  Because Quarto is language‑agnostic, you can also include chunks written in Python or Julia alongside R code, but we will focus on R.

### Quarto basics and YAML

The YAML header sets the global context for your document.  A minimal Quarto file looks like this:

```yaml
---
title: "Global life expectancy report"
author: "Your Name"
date: 2025-11-18
format: html
---
```

The `title` and `date` appear at the top of the rendered document.  The `format` specifies the default output; common values include `html`, `pdf` and `docx`.  You can supply an expanded `format:` field to customize specific options or produce multiple outputs:

```yaml
format:
  html:
    toc: true        # include a table of contents
    code-fold: true  # make code blocks collapsible
  pdf: default
  docx: default
```

With this configuration Quarto will generate HTML, PDF and Word files when you render the document.  If you render from RStudio the appropriate viewers open automatically.  To render from the command line you can run `quarto render my-document.qmd` or use `quarto::quarto_render()` inside R; both commands support an `output_format` argument that overrides the YAML specification.

### Code chunks and options

Code chunks are where your analysis happens.  Each chunk can be given a label so that you can reference its output (e.g., `#| label: setup`).  Chunk **options** control whether code is executed and how results appear.  Here are the most commonly used options:

* `eval: false` prevents the code from running.  Use this to display example code without executing it.
* `include: false` runs the code but hides both the code and its results; this is useful for setup tasks like loading packages or reading data.
* `echo: false` hides the code but shows the results.  Use this when writing reports for people who care about the conclusions but not the implementation.
* `message: false` and `warning: false` suppress messages and warnings from appearing in the output.
* `results: hide` hides printed output; `fig-show: hide` hides plots.

You can set these options on a per‑chunk basis by adding `#| option: value` lines immediately below the chunk header.  For example:

```{r}
#| label: scatterplot-gapminder
#| echo: false
#| warning: false
library(gapminder)
library(ggplot2)

ggplot(gapminder, aes(gdpPercap, lifeExp, color = continent)) +
  geom_point(alpha = 0.6) +
  scale_x_log10() +
  labs(x = "GDP per capita (log scale)",
       y = "Life expectancy (years)",
       color = "Continent")
```

If many chunks share the same settings, set global chunk options in the YAML header under an `execute` field:

```yaml
execute:
  echo: false
  message: false
  warning: false
```

This hides code and suppresses messages by default; you can override these defaults in individual chunks by explicitly setting `echo: true` or similar.

### Rendering to multiple formats

Quarto can produce a variety of outputs beyond static HTML.  In the YAML header you can specify multiple formats under the `format:` field.  Each format can have its own options.  For example, to add a floating table of contents to the HTML version while keeping the PDF and Word versions default:

```yaml
format:
  html:
    toc: true
    toc_float: true
  pdf: default
  docx: default
```

When you render the document, Quarto will create three files: `.html`, `.pdf` and `.docx`.  If you only want to render one format temporarily, supply the `output_format` argument to `quarto_render()`, e.g.,

```{r}
#| eval: false
quarto::quarto_render("my-report.qmd", output_format = "docx")
```

Another useful option for HTML outputs is `code-fold: true`, which hides code by default but allows readers to expand it on demand.  Remember that PDF output requires a LaTeX installation; RStudio will prompt you to install TinyTeX if you don’t have it.

### Inline code and narrative

One of the strengths of literate programming is the ability to embed code directly within text.  Inline code is enclosed in backticks preceded by `r`, like ``r nrow(gapminder)``.  When the document is rendered, the expression is evaluated and its result inserted into your prose.  This ensures that statements about your data remain accurate as your analysis evolves.  To control the formatting of numbers you can define a helper function, such as:

```{r}
#| echo: false
comma = function(x) format(x, digits = 1, big.mark = ",")
```

Then you could write, “The gapminder dataset contains ``r comma(nrow(gapminder))`` observations across ``r length(unique(gapminder$year))`` years.”

### Communicating your findings

Once your analysis is reproducible, you need to communicate it effectively.  A good graphic tells a story.  Start with *labels*: use `labs()` to add clear axis titles, legend titles, a *main title* that summarizes the key finding and an optional *subtitle* and *caption*.  Avoid titles that merely restate the variables (“A scatterplot of GDP vs. life expectancy”) and instead summarize the insight (“Life expectancy increases with wealth, but Europe leads at every income level”).  Use the subtitle for additional context and the caption to cite data sources or provide notes.  When labelling axes and legends, replace short variable names with descriptive phrases and include units (e.g., “GDP per capita (USD)”).  Thoughtful labels help your audience understand your plots without reading your code.

Beyond labels, consider *annotations*, *scales* and *themes*.  Annotations (e.g., `geom_text()`, `geom_label()`) can highlight notable observations or annotate clusters.  Adjust scales to control axis ticks, breaks and legends; the `scales` package provides functions like `label_comma()` for nicely formatted numbers.  Themes control the non‑data elements of the plot: fonts, backgrounds and layout.  The default `theme_minimal()` is a good starting point; use `theme()` to adjust specific elements such as the position of titles and captions.  For multi‑plot layouts, packages like `patchwork` let you combine multiple ggplots and add common titles and captions.

### Key take‑aways

* **Combine code and narrative.**  Use Quarto to weave your code and prose together so that anyone can reproduce your results and understand your reasoning.
* **Structure your document with YAML.**  The YAML header sets the title, date and output formats and can contain default execution options.  Expand the `format:` field to customise each output and render to multiple formats.
* **Master chunk options.**  Use options like `eval`, `include`, `echo`, `message` and `warning` to control what is executed and displayed.  Set global defaults in the YAML and override them locally when necessary.
* **Use inline code for dynamic text.**  Insert the results of R expressions into your prose with inline code and define helper functions to format numbers cleanly.
* **Communicate clearly.**  Titles should summarize the main insight; subtitles and captions provide context and sources; axis labels and legends should be descriptive and include units.  Use annotations, scales and themes to guide your audience’s attention and craft expository graphics.

<!-- ## In‑class activity -->

<!-- The in‑class activity for this session asks students to create a reproducible Quarto report using the **gapminder** dataset from the `gapminder` package.  They will practise setting up a Quarto document, customizing the YAML header, using chunk options to control execution and output, and designing clear graphics.  See the separate assignment description for details. -->

<!-- --- -->

<!-- **Suggested reading for Session 21:**  To prepare for the next class on exploratory data analysis (EDA), read **Chapter 10 “Exploratory data analysis”** from *R for Data Science* (2nd ed.).  The chapter frames EDA as an iterative cycle where you generate questions, search for answers using visualization and transformation, and refine your questions based on what you learn【728943193213359†L111-L126】.  EDA is not a rigid procedure but a creative process; ask lots of questions to uncover patterns, anomalies and relationships.  The authors highlight two overarching questions to guide your exploration: “What type of variation occurs within my variables?” and “What type of covariation occurs between my variables?”【728943193213359†L167-L176】.  Read about how to visualise variation (typical and unusual values) and covariation between variables, and consider the follow‑up questions suggested for each plot.  This reading will lay the foundation for systematically exploring datasets in the next session. -->