# Session 20: Reproducible research and communication

## Objectives

- **Appreciate the value of reproducible research.** Combining code, results, and prose in a single document improves transparency, facilitates collaboration, and serves as a lab notebook. Quarto integrates the lessons of the R Markdown ecosystem and is designed for analysts, collaborators, and decision-makers alike.
- **Learn the basics of Quarto documents.** Create and render a `.qmd` file with a YAML header and interleaved code and narrative. Understand global settings in the YAML, chunk labels, and what actually happens when a label is reused by accident.
- **Use chunk options effectively.** Control evaluation, code display, warnings, and messages with chunk options (`eval`, `include`, `echo`, `message`, `warning`). Learn how inline code reports dynamic values inside your prose, and what happens when that inline code itself fails.
- **Distinguish caching from freezing.** Recognize `cache: true` (a local, per-chunk performance cache) as a different mechanism from `execute: freeze` (the project-wide, version-controlled cache this very book relies on), and know which one solves which problem.
- **Communicate your findings clearly.** Turn exploratory graphics into expository graphics. Write titles that summarize the finding rather than describing the plot; add subtitles and captions; label axes and legends with units; and zoom into a plot honestly, without silently changing what a fitted trend line is actually fit to.

## Notes

```{r}
#| message: false
#| warning: false
library(tidyverse)
```

### Why reproducible research?

In an exploratory notebook, it's easy to lose track of which code produced which result. Reproducible research solves this by knitting analysis and narrative together, so that anyone, including future you, can rerun the whole thing and get the same output. This is sometimes called literate programming: code interspersed with human-readable explanation. Quarto, the modern successor to R Markdown, is the tool this course (and this very book) uses for it. A `.qmd` file has three parts: a YAML header (metadata and settings, delimited by `---`), narrative text in Markdown, and code chunks that Quarto actually executes when rendering.

### Quarto basics and YAML

A minimal Quarto document's header looks like this:

```yaml
---
title: "Global life expectancy report"
author: "Your Name"
date: 2025-11-18
format: html
---
```

`format` sets the default output; common values are `html`, `pdf`, and `docx`. An expanded `format:` field customizes options, or produces more than one output at once:

```yaml
format:
  html:
    toc: true            # table of contents
    code-fold: true      # collapsible code blocks
    embed-resources: true # bundle images/CSS into one standalone .html file
  pdf: default
  docx: default
```

`embed-resources` is worth knowing about specifically: without it, an HTML file's images and styling live in separate files alongside it, which breaks the moment you email just the `.html` on its own; with it, everything is bundled into one self-contained file. Rendering happens with `quarto render my-document.qmd` from the command line, or `quarto::quarto_render()` from R (both support an `output_format` argument that overrides the YAML). A `params` field in the YAML can declare inputs (accessed as `params$name` inside the document), turning a single `.qmd` file into a template you can re-render for different inputs without editing the file itself.

### Code chunks and options

Each code chunk can carry a label (`#| label: setup`) for cross-referencing, and options that control execution and display: `eval: false` shows code without running it, `include: false` runs code but hides both it and its output (handy for setup chunks), `echo: false` hides code but shows results, and `message: false`/`warning: false` suppress exactly what they say. Options go on `#| option: value` lines right after the opening fence.

```{r}
#| label: scatterplot-gapminder
#| echo: false
#| warning: false
library(gapminder)

ggplot(gapminder, aes(gdpPercap, lifeExp, color = continent)) +
  geom_point(alpha = 0.6) +
  scale_x_log10() +
  labs(
    x = "GDP per capita (log scale)",
    y = "Life expectancy (years)",
    color = "Continent"
  )
```

Chunk labels have to be unique across the whole document; reusing one is a real, render-stopping mistake, not a cosmetic issue (see Example 20.2). Setting shared options once under `execute:` in the YAML (`echo: false`, `message: false`, `warning: false`, exactly the settings this book itself uses) saves repeating them in every chunk; an individual chunk can still override the default locally.

### Tables with `knitr::kable()`

A data frame printed directly inside a chunk shows console-style output; `knitr::kable()` instead formats it as a proper Markdown (and, on render, HTML or PDF) table.

```{r}
mtcars |>
  select(mpg, cyl, disp) |>
  head() |>
  knitr::kable()
```

For anything beyond a simple table (styling, merged cells, interactivity), packages like `gt`, `kableExtra`, and `reactable` build on the same idea with considerably more control.

### Rendering to multiple formats

Quarto renders one `.qmd` file into as many formats as the YAML's `format:` field lists, each with its own options:

```yaml
format:
  html:
    toc: true
    toc-float: true
  pdf: default
  docx: default
```

Rendering this document produces `.html`, `.pdf`, and `.docx` files in one call. To render just one format on demand rather than every configured one, pass `output_format` to `quarto_render()`:

```{r}
#| eval: false
quarto::quarto_render("my-report.qmd", output_format = "docx")
```

PDF output specifically needs a LaTeX installation; RStudio offers to install a minimal one (TinyTeX) automatically the first time you render to PDF without one.

### Inline code and narrative

Inline code, an R expression wrapped in a special inline-code span, evaluates that expression and inserts its result directly into your prose when the document renders, which keeps statements about your data accurate as the underlying analysis changes. A small helper function keeps inline number formatting consistent:

```{r}
#| include: false
comma <- function(x) format(x, digits = 1, big.mark = ",")
```

Writing two such inline expressions in a sentence, one calling `comma(nrow(gapminder))` and the other calling `length(unique(gapminder$year))`, each wrapped in inline-code syntax, renders as: The gapminder dataset contains `r comma(nrow(gapminder))` observations across `r length(unique(gapminder$year))` years. If the expression inside an inline code span itself errors, that failure isn't quietly swallowed either; the whole render stops (see Example 20.3), the same as a code chunk failing would.

### Caching versus freezing

Two different mechanisms both promise "don't rerun code that hasn't changed," and they solve different problems. Chunk-level `cache: true` is a local performance optimization: knitr stores a chunk's results in a hidden `_cache` folder on your own machine, meant to save time while you iterate, not to be committed to version control. Project-wide `execute: freeze: auto`, the setting this book's own `_quarto.yml` actually uses, works differently: it stores each chapter's execution results in a `_freeze` folder that *is* committed to git, so that Continuous Integration (or a collaborator on a different machine) can render the whole book without needing every package this book uses installed, reusing the cached results for any chapter whose source hasn't changed. Confusing the two is an easy mistake (see Example 20.4).

### Communicating your findings

A reproducible analysis still needs to communicate clearly. `labs()` is the starting point: a title, subtitle, caption, and axis/legend labels.

```{r}
lifeexp_2007 <- gapminder |> filter(year == 2007)

ggplot(lifeexp_2007, aes(gdpPercap, lifeExp, color = continent)) +
  geom_point(alpha = 0.6) +
  scale_x_log10(labels = scales::label_dollar()) +
  labs(
    title = "Wealthier countries tend to live longer, but Europe leads at every income level",
    subtitle = "Each point is one country in 2007",
    x = "GDP per capita",
    y = "Life expectancy (years)",
    color = "Continent",
    caption = "Source: gapminder"
  )
```

A title should summarize the finding ("Wealthier countries tend to live longer...") rather than merely describe the plot ("A scatterplot of GDP vs. life expectancy"); the subtitle adds context, and the caption is a natural place for a data source. `scales::label_dollar()` and `scales::label_percent()` format axis breaks the way a reader actually expects to see money or proportions, rather than as raw numbers.

Beyond labels, `annotate()` and `geom_hline()`/`geom_vline()` call out a specific value or region directly on the plot:

```{r}
ggplot(lifeexp_2007, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = mean(lifeexp_2007$lifeExp), linetype = "dashed") +
  annotate("text", x = 500, y = mean(lifeexp_2007$lifeExp) + 2, label = "Global average", hjust = 0) +
  scale_x_log10(labels = scales::label_dollar())
```

`theme_minimal()` (used throughout this book) is a reasonable default for reducing visual clutter; `theme()` adjusts specific elements (title position, legend placement) beyond what a built-in theme covers, and the `patchwork` package combines several separate ggplots into one figure with `+`/`|`/`/`, handy for a multi-panel figure with one shared title.

Zooming into part of a plot deserves special care: `coord_cartesian(ylim = ...)` changes only what's visible, while filtering the data first (or setting a scale's `limits`) actually removes data before anything, including a fitted trend line, is computed, which can silently change the numbers a plot is telling you about (see Example 20.1).

### Fringe cases and common pitfalls

::: {.example title="Example 20.1"}
**Zooming with `coord_cartesian()` and zooming by filtering data are not the same operation, and a fitted line proves it.**

```{r}
set.seed(1)
df <- tibble(x = 1:100, y = x + rnorm(100, 0, 10)) |>
  bind_rows(tibble(x = c(5, 10), y = c(200, 250)))  # two extreme outliers

full_fit <- lm(y ~ x, data = df)
coef(full_fit)

filtered_fit <- lm(y ~ x, data = df |> filter(y <= 120))
coef(filtered_fit)
```

The slope and intercept are meaningfully different (roughly 0.79 versus 1.00) depending on whether the two outliers were included when the line was fit, even though both versions might be *displayed* zoomed into the exact same visible range. `coord_cartesian(ylim = c(0, 120))` would zoom the plot's viewport into that range while still fitting `geom_smooth()` on every point, outliers included; filtering the data down to `y <= 120` first, or setting a scale's `limits` (which drops out-of-range values the same way filtering does), actually removes the outliers before any statistic is computed. Visually, both can look like "the same zoomed-in plot"; only one of them is telling you the truth about a trend fit to the *complete* data.
:::

::: {.example title="Example 20.2"}
**A duplicate chunk label doesn't just cause a cosmetic warning; it stops the whole render.**

Two chunks in the same document both labeled `#| label: setup` produce this real error:

```
Error in parse_block(g[-1], g[1], params.src, markdown_mode) :
  Duplicate chunk label 'setup', which has been used for the chunk:
1 + 1
Execution halted
```

Quarto uses chunk labels for cross-references and for naming generated figure files, so it needs every label in a document to be unique, and it refuses to guess which "setup" chunk you actually meant. This is exactly the kind of error most likely to show up after copying a chunk (setup code, a common summary pattern) from an earlier session's notes into a new one without renaming its label. If a render fails with "Duplicate chunk label," search the document for that exact label; there will be at least two.
:::

::: {.example title="Example 20.3"}
**An error inside inline code halts the render too, the same as an error inside a full code chunk.**

Writing a sentence that references an object that doesn't exist through inline R code, such as an undefined `nonexistent_variable_xyz`, produces a real, render-halting failure the moment Quarto tries to evaluate it:

```
Error:
! object 'nonexistent_variable_xyz' not found
Quitting from report.qmd:42-42
Execution halted
```

(This example describes the error rather than triggering it live, since doing so would halt the render of this very chapter, which is exactly the point being illustrated.) It can be tempting to think of inline code as "just text with a little bit of R sprinkled in," safer somehow than a full chunk, but Quarto evaluates it exactly the same way, and a failure there is exactly as fatal to the render as a failure inside a chunk. This is arguably the correct behavior (a broken inline value silently rendering as nothing, or as a stale cached number, would be far worse than a loud failure), but it does mean a single typo in a sentence can prevent an entire document, tables and plots included, from rendering at all.
:::

::: {.example title="Example 20.4"}
**`cache: true` and `execute: freeze` solve similar-sounding problems in genuinely different ways, and mixing them up causes real confusion.**

`cache: true` on an individual chunk tells knitr to save that chunk's results locally (in a `_cache` folder next to the document) so that re-rendering the same document on the same machine can skip recomputing it, purely as a speed optimization for you, right now, while you iterate. It is not meant to be committed to git, and by default it does not know to invalidate itself if a chunk depends on an external file that changed (you'd need `cache.extra = file.mtime("data/some_file.csv")` to teach it that).

This book's own `_quarto.yml` uses something different: `execute: freeze: auto`, which stores each chapter's full execution results in a `_freeze` folder that *is* committed to git. That's precisely why the GitHub Actions workflow that builds and publishes this book doesn't need to install `arrow`, `duckdb`, or every other package used somewhere in these twenty-plus chapters just to rebuild the one chapter you actually edited; unchanged chapters reuse their committed `_freeze` results, and only a chapter whose source actually changed gets re-executed. `cache: true` speeds up your own local iteration on one document; `execute: freeze` is what makes a whole multi-chapter book reproducible for anyone (or any CI system) that clones the repository.
:::

### Recap

| **Term** | **Definition** |
|-------------------|-------------------------------------------------------------|
| **YAML header** | The `---`-delimited metadata block at the top of a `.qmd` file: title, author, format, and execution defaults. |
| **Chunk label** | A name (`#| label: ...`) identifying a code chunk; must be unique in the document, or the render fails. |
| **`embed-resources`** | An HTML output option that bundles images and styling into one self-contained file. |
| **`params`** | YAML-declared inputs (accessed as `params$name`), turning a `.qmd` into a re-render-able template. |
| **`knitr::kable()`** | Formats a data frame as a proper table in the rendered output, instead of console-style printed text. |
| **`cache: true`** | A local, per-chunk performance cache (not meant for git); doesn't auto-invalidate on an external file change without `cache.extra`. |
| **`execute: freeze`** | A project-wide, git-committed cache of each document's execution results, used for CI/collaborator reproducibility. |
| **Inline code** | An R expression wrapped in inline-code syntax inside prose; evaluated at render time, and just as capable of halting the render as a full chunk. |
| **`coord_cartesian()`** | Zooms a plot's visible range without discarding any underlying data, so fitted statistics (like `geom_smooth()`) stay computed on the full dataset. |

### Check your understanding

::: {.callout-note collapse="false" title="Problems"}
1. What are the three components of a `.qmd` file, and what is each one responsible for?
2. A report needs to be re-generated once per region, with only the region name changing each time. Which Quarto YAML feature is built for exactly this, and how would you access the region name inside the document?
3. Two chunks in the same document are both labeled `analysis`. What happens when you try to render it, and why does Quarto insist on unique labels in the first place?
4. You zoom into a scatterplot with a fitted trend line by filtering out points above a threshold, rather than using `coord_cartesian()`. Explain how this could change the reported trend, not just how much of the plot is visible.
5. A colleague says they added `cache: true` to a slow chunk that reads an external CSV, and now the chunk doesn't rerun even after they edited that CSV. What is actually going on, and how is this different from how this book's own chapters stay up to date?
:::

::: {.callout-tip collapse="true" title="Solutions"}
1. A YAML header (metadata and default settings, such as title and output format), narrative text in Markdown (headings, paragraphs, lists), and code chunks (executable code whose output is inserted into the rendered document).

2. The `params` field in the YAML header declares an input, accessed inside the document as `params$name` (for example, `params$region`); re-rendering with a different value passed for that parameter regenerates the report for a different region without editing the `.qmd` file's actual content.

3. Quarto refuses to render and reports a "Duplicate chunk label" error naming the reused label, because labels are used to identify chunks for cross-referencing and to name generated figure files, and Quarto has no way to guess which of the two identically labeled chunks you actually meant in either context.

4. Filtering the data before fitting a trend line removes those points from the calculation entirely, not just from the visible plot area, so the fitted line reflects a different (smaller) dataset than the one being discussed. `coord_cartesian()` only changes the visible viewport; every point, including ones now outside the visible range, is still included when the trend line itself is computed.

5. `cache: true` is a local performance cache that does not automatically know to invalidate itself when a chunk's underlying dependency (like the CSV file) changes; it only detects changes to the chunk's own code, unless the colleague explicitly adds something like `cache.extra = file.mtime("data.csv")` to make the cache dependent on that file too. This book's own reproducibility instead relies on `execute: freeze`, a different, project-wide, git-committed mechanism, and it faces an analogous risk: a frozen chapter's cached results also won't update on their own if only a referenced data file changes and the chapter's own `.qmd` source doesn't, which is why deleting the relevant `_freeze` entry (or re-rendering explicitly) is sometimes necessary after changing shared data.
:::
